Dynamic modes of Notch transcription hubs conferring memory and stochastic activation revealed by live imaging the co-activator Mastermind

Version 1

Abstract

Summary

Developmental programming involves the accurate conversion of signaling levels and dynamics to transcriptional outputs. The transcriptional relay in the Notch pathway relies on nuclear complexes containing the coactivator Mastermind (Mam). By tracking these complexes in real time, we reveal that they promote formation of a dynamic transcription hub in Notch-ON nuclei which concentrates key factors including Mediator CDK module. The composition of the hub is labile and persists after Notch withdrawal conferring a memory that enables rapid reformation. Surprisingly, only a third of Notch-ON hubs progress to a state with nascent transcription, that correlates with Polymerase II and core Mediator recruitment. The probability is increased by a second signal. The discovery that target-gene transcription is probabilistic has far-reaching implications because it implies that stochastic differences in Notch pathway output can arise downstream of receptor activation.

Introduction

Cells face the challenge of transmitting information accurately, so that cell-surface signals are translated into correct transcriptional responses, and how this is achieved mechanistically remains a major question. Notch is a key signalling pathway that leads to gene activation when ligand and receptor engage upon cell contact 1–3. The physical interaction brings about a conformational change that permits proteolytic cleavage and release of Notch Intra Cellular Domain (NICD) 4–6. This moiety forms a complex with CSL (CBF-1/RBPJ-κ in mammals, Suppressor of Hairless in Drosophila and LAG-1 in C. elegans), a transcription factor that binds to specific DNA motifs, and Mastermind (Mam) a co-activator 7–9. This tripartite activation complex promotes transcription from the target genes where it is recruited. The sites of recruitment differ according to the cellular context, resulting in different transcriptional outcomes and suggesting that other factors are important in preparing the targets for activation 10. In addition, release of NICD brings about rapid and robust transcriptional responses within minutes, raising the question how the molecules of cleaved Notch achieve this so efficiently 11–13.

Regulation of transcription must be tightly controlled in space, time, and genomic location 14,15. Many different studies report that sequence-specific transcription factors, key coactivators, and RNA polymerase II (PolII) itself undergo dynamic clustering within a nucleus 16–19. Clustering appears to be mediated by a combination of specific structure-mediated interactions (e.g. DNA-binding, protein-protein interactions) and multivalent interactions among intrinsically disordered regions (IDRs) present in most transcription factors 20,21. In this way, transcription is regulated by the formation of functionally specialized local protein microenvironments or transcription “hubs” associated with target enhancers 22. In some cases, these have the properties of ‘condensates’ whose formation and dissolution has been explained by the process of phase separation 23. As the resulting assembly is non-stoichiometric, it may enable a small number of transcription factor molecules to drive productive transcription. Such a mechanism could thus explain how NICD, whose nuclear levels are frequently below the level of detection in vivo, can successfully promote robust-target gene transcription 24. Indeed, all members of the Notch activator complex contain unstructured regions that could contribute to the assembly of a hub.

The formation of the tripartite Notch activator complex involves a conserved helix in the N-terminal region of Mam proteins which is responsible for the direct interactions with CSL and NICD 8,9. The remainder of the large Mam proteins are poorly conserved and appear to be predominantly unstructured albeit to have potential roles in recruiting other cofactors 25,26. For example, there is evidence that human MAML1 interacts with the histone acetyl transferase CBP/p300, which is present at Notch regulated enhancers in genome-wide studies 26–28, and whose recruitment is implicated in activating some targets 29,30. The C-terminal portion of MAML1 is also suggested to recruit CDK8, the enzymatic core of the Mediator kinase module 31–33. What role these factors play in the recruitment, dynamics and assembly of functional Notch transcription assemblies, and whether these acquire hub-like properties is unclear.

Live imaging of endogenously tagged proteins offers a non-invasive approach to probe the assembly and composition of transcription hubs in vivo. Using this strategy, we have previously shown that CSL is recruited in a very dynamic manner to a target genomic locus in vivo 34. However, as CSL exists in both corepressor and co-activator complexes 35, the extent that these dynamics reflect the characteristics of the activation complex remains to be established. Here we incorporated fluorescent tags into the endogenous Mam protein to investigate the behaviours of this Notch nuclear co-activator in vivo, in combination with a method for live-imaging of a target genomic locus that responds robustly to Notch activation. The emerging model is that Notch activity leads to the formation of ‘transcription hubs’ that exist in different states. When Mam is present, key components of the transcriptional machinery become locally concentrated but, surprisingly, in the absence of synergising factors, the conversion to productive transcription only occurs stochastically. In addition, the open chromatin state that is generated decays slowly after Notch withdrawal, providing a memory that enables a more rapid response to a subsequent round of Notch activation.

Results

Dynamics of Mastermind recruitment in relation to its partner CSL

The Mam coactivator is an integral part of the Notch transcription complex 8,9,30,36. To track Mam dynamic behaviours in vivo we inserted GFP or Halo into the N-terminus of the endogenous Mam using CRISPR-Cas9 genome editing 34. The resulting flies are homozygous viable with no evident phenotypes, indicating that the tagged Mam proteins are fully functional. We first set out to compare the recruitment and dynamics of CSL and Mam at a target locus in Notch-OFF and Notch-ON conditions, taking advantage of Drosophila salivary glands, where the polytene (multiple copy) chromosomes aid detection of chromatin-associated complexes. We used the Int/ParB system, where fluorescently labelled ParB proteins bind to inserted Int sequences, to detect the well-characterised Enhancer of split complex [E(spl)-C], which contains multiple Notch-regulated genes34. Because of the aligned copies of the genome in the polytene chromosomes, the target genomic locus appears as an easily distinguished fluorescent “band” in each nucleus during live imaging (Figure 1A).

Enrichment and dynamics of Mastermind at E(spl)-C.

(A) Schematic overview of live imaging system used. Drosophila salivary glands have large nuclei with polytene chromosomes that facilitate imaging. E(spl)-C locus is labelled using Int (orange)/ParB (red) system. Activation complexes, [CSL (green), NICD (purple) and Mastermind (magenta)] and co-repression complexes [CSL and Hairless (brown)] are depicted.

(B) Live-imaging of GFP::Mam and GFP::CSL as indicated in relation to E(spl)-C marked by Int-ParB (magenta) in nuclei from Notch-OFF (1151Gal4; UAS-LacZ) and Notch-ON (1151Gal4; UAS-NΔECD) salivary glands. CSL and Mam are enriched at E(spl)-C in Notch-ON but not Notch-OFF cells. Average enrichment: each pixel represents average enrichment of all aligned images, centred on E(spl)-C locus. Enrichment profile: mean enrichment, with SEM, plotted on x-axis relative to position, y axis, centred on E(spl)-C (0). Grey area indicates region used for max enrichment. Max enrichment: mean of 10 pixels centred on E(spl)-C (Figure S1A) (Mam OFF, n=32, Mam OFF, n = 30; CSL OFF, n = 45, CSL ON, n = 28, for p-values see Table S3). Box encompasses range between 0.25 and 0.75 quantile, whiskers extend to furthest points not considered outliers, bar marks median, cross marks mean and each dot is the value for one nucleus. Scale bars represent 5um.

(C) Dynamics of CSL::GFP (green), GFP::Mam (magenta) and Hairless::GFP (brown) at E(spl)-C in Notch-ON cells measured by FRAP. Dots indicate 50% recovery. Legend summarizes numbers of nuclei (n) and time to 50% recovery (t1/2). Error represents the standard error of the mean (SEM). (C’) FRAP analysis of CSL and Mam in cells depleted for Hairless (1151Gal4; UAS-Hairless RNAi) controls from C are included for comparison.

(D) Trajectory density at E(spl)-C relative to whole nucleus from SPT of CSL::Halo (green) and Halo::Mam (pink) in Notch-OFF and Notch-ON cells. Both CSL and Mam are significantly enriched in Notch-ON. Box plots as in (B). CSL-OFF, n = 5, CSL-OFF, n = 7; Mam-OFF, n = 5, Mam-ON, n = 13; for p-values see Table S3.

(E, E’) Survival curves depicting duration of trajectories in SPT of the indicated Halo fusion proteins in whole nuclei (E) and at E(spl)-C (E’), in Notch-ON cells. Whole nuclei H2B::Halo trajectories are included for comparison in both graphs. For number of trajectories, see Figure S1D. Error bars represent 95% confidence intervals, obtained from bootstrapping with 100 resampled datasets 96.

See also Supplementary Figure S1.

The Notch pathway is normally inactive in salivary glands, providing a baseline Notch-OFF condition. This was converted to Notch-ON by the expression of a constitutively active form of Notch, NΔECD, using the GAL4-UAS system which allows tissue-specific and temporal control. Because it lacks the extracellular domain, NΔECD is constitutively cleaved by gamma-secretase to release NICD, mimicking ligand induced activation 37–39. Comparing the localization of GFP::Mam in Notch-OFF and Notch-ON conditions, it was immediately evident that Mam was robustly recruited to E(spl)-C in Notch-ON conditions in a similar manner to CSL 34. In Notch-OFF conditions, both proteins were diffuse throughout the nucleus, with a low level of CSL, but not Mam, present at E(spl)-C. In Notch-ON conditions, strong enrichment of both Mam and CSL was consistently detected around E(spl)-C (Figure 1B).

To compare the dynamics of Mam and CSL at E(spl)-C in Notch-ON conditions, we performed Fluorescence Recovery After Photobleaching (FRAP) focussed on the region defined by the locus-tag. Unlike CSL, which had a rapid recovery (t1/2=9 secs), Mam exhibited much slower dynamics (t1/2=40 secs) and failed to fully recover over the time-course of the experiment (Figure 1C). Slower recovery can arise from higher proportion of bound molecules, longer residence times and/or slower diffusion coefficients40. The results therefore suggest that the Mam-containing activation complexes have different properties from the majority of CSL complexes.

The other main partner for CSL in Drosophila is the corepressor Hairless. In FRAP experiments Hairless has a fast recovery, with a profile close to that of CSL (Figure 1C), consistent with a significant fraction of CSL being complexed with Hairless even in Notch-ON conditions34. The difference in dynamics between CSL and Mam could therefore be explained by the former being involved in two different complexes with different dynamics. To test this, we depleted Hairless using RNAi-mediated knock-down and measured the effects on enrichment and dynamics of CSL and Mam. As expected, Mam recruitment levels and the dynamics measured by FRAP were unchanged (Figure 1C’, S1B). In contrast, CSL levels and recruitment were reduced and its FRAP recovery was slowed, albeit not to the extent that it recapitulated the Mam profile (Figure 1C’, S1C). Together the results indicate that the activation complexes, containing CSL and Mam, have slower dynamics than the repressor complexes, containing CSL and Hairless, and that the recovery of CSL reflects its participation in the two types of complexes.

To further investigate the dynamics of Mam and CSL complexes, we performed single particle tracking (SPT) by sparse labelling of endogenous Halo::Mam and Halo::CSL in live tissue 41,42. Gaussian fitting-based localisation and Multiple hypothesis tracking were used for detection and tracking of single particles within the nucleus with a ~20nm precision34. Using a Bayesian treatment of Hidden Markov Model, vbSPT 43, trajectories were assigned into 2 states, defined by a Brownian motion diffusion coefficient, that correspond to bound (diffusion coefficient 0.01µm2/s) versus more freely diffusing complexes (diffusion coefficient >0.25µm2/s) (Figure S1D). A greater proportion (53%) of Mam complexes were in the bound state than CSL complexes (39%), consistent with the differences between their FRAP curves. We also analysed the density of particle trajectories in relation to the E(spl)-C locus in Notch-OFF and Notch-ON conditions. In comparison to their average distribution across the nucleus, both CSL and Mam trajectories were significantly enriched in a region of approximately 0.5 µm around the target locus in Notch-ON conditions, reflecting robust Notch dependant recruitment to this gene complex.

To assess whether Mam complexes have longer residence times once recruited to the chromatin, we analyzed the duration of trajectories for Mam, CSL and Hairless. Long trajectories correlate to bound complexes, because faster moving particles are rapidly lost from the field of view, and the length of time they are detectable is an indication of relative residence times. There were clear differences between the trajectory durations for Mam, CSL and Hairless. Mam trajectories had the longest durations of up to 15 secs, Hairless trajectories were the shortest (up to 5-7 secs) and CSL trajectories were intermediate (up to 10 secs) (Figure 1E). The differences were recapitulated when only the trajectories in the region around E(spl)-C were analysed (Figure 1E’). The residences are likely an underestimation because bleaching and other technical limitations also affect the track durations44. Nevertheless, these data confirm that Mam-containing complexes have on average longer residence times than other CSL complexes which, together with the higher proportion of bound molecules overall, explains the slower recovery dynamics measured by FRAP.

The fact that CSL dynamics, measured by FRAP and SPT, are intermediate between Hairless and Mam fit well with it being present in two types of complexes [co-activator (Mam) and corepressor (Hairless)]. However, whether the contribution from CSL co-repressor complexes can fully explain all the observed differences between CSL and Mam is not fully clear. First, when we depleted Hairless, so that the majority of CSL present would be in co-activator complexes, CSL FRAP recovery curves were still substantially different from those of Mam. Second, none of the CSL trajectories had a duration approximating those of the longest Mam trajectories, despite that over 50,000 CSL trajectories were tracked (compared to 14,000 Mam, Figure S1D). It is possible therefore that, once recruited, Mam can be retained at target loci independently of CSL by interactions with other factors so that it resides for longer.

Hub-like properties of CSL-Mastermind complexes in Notch active cells

The enrichment of CSL and Mam around E(spl)-C that we detect by live-imaging is not unexpected because this locus has multiple genes containing CSL binding-motifs 34,45. However, the diffuse enrichment was not maintained when the tissues were fixed, a characteristic reported for proteins present in condensate-like hubs 46 (Figure S2A). The localized concentration of exchanging CSL and Mam complexes around E(spl)-C in Notch-ON nuclei may therefore have properties of a transcription hub, with some of the recruitment being reliant on weak interactions mediated by low-complexity regions 47–49.

First, we asked to what extent CSL recruitment was correlated with the number of CSL binding motifs. To do so, we took advantage of fly strains containing multiple copies of CSL motifs inserted at an ectopic position in the genome 50 and compared the recruitment with insertions containing 12 or 48 CSL motifs. In Notch-ON conditions, these insertions were sufficient to generate an ectopic band of CSL recruitment similar to the native E(spl)-C (Figure 2A). Remarkably the amounts of CSL recruited to loci with 12 and 48 ectopic sites were almost identical. Thus, there is not a direct correlation between the number of motifs and the amount of CSL recruited under these conditions. Nor does it appear that the amounts of CSL complexes are a limiting factor as there was no decrease in recruitment at the endogenous E(spl)-C locus, even in nuclei with a 48 CSL-motif insertion.

Hub-like properties of recruited complexes in Notch-ON nuclei.

(A) Representative images of GFP::CSL in Notch-ON nuclei containing an ectopic array of 12 or 48 CSL binding sites. GFP::CSL is recruited to E(spl)-C (yellow) and the ectopic array (blue). Average enrichment, Enrichment profile as in Figure 1B; (ectopic array 12, n=45, ectopic array 48, n = 40; E(spl)-C 12, n=45, E(spl)-C 48, n=40).

(B) Domain organisation of the indicated proteins are diagrammed above the prediction scores (IUPRED) of protein disorder for each. Regions tested as IDRs are indicated.

(C) Average enrichments and enrichment profiles of NICD-PEST IDR (Notch-OFF, n = 62, Notch-ON, n = 67) and CSL-nt IDR (Wilcoxon rank sum test: Notch-OFF, n = 8, Notch-ON, n = 40) at E(spl)-C in Notch-OFF and Notch-ON cells.

(D) Average enrichments and enrichment profiles (as in Figure 1B) of GFP::Mam at E(spl)-C in nuclei expressing intact (Nact) and IDR deleted (Nact ΔIDR) Notch constructs (Nact, n = 58, Nact ΔIDR, n = 55). p-values in Table T3

(E) Diagram illustrates concentric ring analysis of single particle trajectories at increasing distances from E(spl)-C. Graphs plot proportions of bound (dark shading) and diffusive (light shading) particles of Halo::Mam or Halo::CSL within each ring. Dashed lines indicate nuclear proportions of each population as indicated.

See also Supplementary Figure S2

Second, we questioned whether the non-stochiometric recruitment of activator complexes to loci with CSL motifs might involve additional weak protein-protein interactions between some components, as observed in several transcription hubs where intrinsically disordered protein domains play a part 20,51. We used Intrinsically Disordered Region (IDR) prediction algorithms 52 to identify IDRs within NICD, CSL, Mam and Mediator1 (Med1, part of the mediator complex) (Figure 2B) and generated transgenic flies where each IDR was expressed as a fusion with GFP. The recruitment of the IDR::GFP fusions to E(spl)-C was then measured in Notch-ON and OFF conditions (Figure 2C). Of the IDRs tested, the C-terminal region from NICD was the most strongly enriched at E(spl)-C in Notch-ON conditions. A very low level of enrichment was evident for a shorter NICD fragment, that lacked the C-terminal PEST sequences, for the CSL C-terminus and for Med 1 IDR. Surprisingly, little or no enrichments occurred with Mam-IDRs, despite that this large protein has been reported to interact with p300 and other factors (Figure S2B). Notably, Mam-nIDR::GFP fusion was present in droplets, suggesting it can self-associate when present in a high local concentration (Figure S2B, 53). Our results suggest therefore that the IDR in NICD may contribute to the localized enrichments at target loci in Notch-ON cells, potentially in combination with IDRs present in other recruited factors. In support of this hypothesis, deletion of the IDR from NICD led to a reduction in the levels and stability of Mam recruitment to E(spl)-C (Figure 2D).

Third, we reasoned that the presence of a transcription hub, where complexes are retained in the vicinity via protein interactions, should result in local changes in the behaviours of CSL and Mam. After segregating the SPT trajectories according to their diffusion properties as described above (Figure S2D), we analyzed the spatial distribution of the slow and fast populations in relation to the E(spl)-C, defined as the area within 550nm of the locus-tag. Based on the shape and centre of this region, concentric zones were defined at 550nm distances and the proportions of slow- and fast-moving particles in each zone were calculated. The results show that there is an enrichment for slow-diffusing and a depletion of fast-moving particles close to the locus, with these altered properties extending to a region of up to 1μm away (Figure 2E).

Together our data support the model that CSL-Mam complexes are recruited and form a hub of high protein concentrations around the target locus in Notch-ON conditions and suggest that IDR interactions, as well as DNA binding, contribute to their recruitment and retention in a region surrounding the active enhancers.

Mediator CDK module is required for stable Mastermind recruitment

The hub-like properties and slower turnover of Mam complexes compared to CSL suggest that other factors will be involved in their stabilization. We therefore tested the consequences of inhibiting or depleting different factors to distinguish those required for Mam enrichment at E(spl)-C in Notch-ON nuclei. We first asked whether active transcription was required for Mam recruitment, by exposing the tissue to a specific inhibitor of transcription initiation, triptolide 54. No change in Mam enrichment or recovery was detected, arguing that Mam recruitment is not dependent on initiation or RNA production (Figure 3A, B). As previous studies have reported an interaction between Mam and the histone acetylase CBP/p300 30,33,55,56, we next inhibited CBP activity using a potent and selective inhibitor A485 57. Tissues were exposed to A485 for one hour which was sufficient to severely reduce the levels of H3K27Ac and E(spl)m3 transcription, indicating that the treatment was effective (Figure 3C, S3A-B). Surprisingly however, there was no change in the recruitment of Mam in these conditions (Figure 3C). We also depleted CBP by RNAi and, as with the drug treatment, Mam recruitment was unchanged (Figure 3D). Based on these results, we conclude that CBP is not essential for the recruitment of Mam complexes to the hub formed at the E(spl)-C locus.

Mam enrichment requires Mediator CDK module but is independent of transcription and CBP/p300.

(A-B) GFP::Mam recruitment levels and dynamics at E(spl)-C in Notch-ON tissues treated with triptolide or DMSO as a control. (A) Average and Max enrichment as in Figure 1B, (DMSO n = 49, triptolide, n = 36). (B) FRAP recovery curves, plotted as in Figure 1C.

(C–F) Average enrichment and Max enrichment of GFP::Mam and GFP::CSL, as indicated, at E(spl)-C in Notch-ON control and treated tissues. (C-D) No change in recruitment following inhibition (C, A485) or genetic knockdown (D, RNAi) of CBP/p300 compared to control, DMSO and yellow RNAi (yRi) respectively. (DMSO, n = 47, A485, n = 62; yRNAi, n = 26, CBP/p300 RNAi, n = 31). (E, F) Reduced recruitment of GFP::Mam (magenta) but not CSL::GFP (green) following (E) genetic knockdown of Med13 and (F) inhibition of CDK8 with Senexin B. (E: Mam control, n = 30, Mam Med13Ri, n = 45; CSL control, n=92, CSL Med13Ri, n = 36) and (F: Mam ctrl n = 27, Mam Senexin, n = 43; CSL ctrl, n = 34, CSL Senexin, n = 37).

(G) Genetic knockdown of Cdk8 results in decreased Mam recruitment compared to controls (ctrl, n=33, Cdk8 Ri, n = 32). Box plots in A-F as in Figure 1B. For p-values see Table S3.

(H) Average enrichment and max enrichment of Med13::YFP at E(spl)-C in Notch-OFF, n = 28 and Notch-ON, n = 31; for p-values see Table S3.

See also Supplementary Figure S3.

As several studies have suggested that the Mediator CDK module, containing Med12, Med 13, CDK8 and CycC, influences Notch dependent transcription, we next focused on the role of this complex 31,32,58. First, we depleted Med13 and CDK8 using validated RNAi lines 59,60. Med13 depletion led to a loss of E(spl)m3 transcription, confirming its role (Figure S3D). In these conditions we detected a substantial reduction in Mam recruitment to E(spl)-C in Notch-ON cells, suggesting the CDK module plays a role in retaining Mam at target sites (Figure 3E, G). Second, we incubated the tissues in Senexin B or Senexin A, two drugs that target CDK8 activity 61,62 for 1 hour prior to imaging. In both cases the treatment was sufficient to significantly impair Mam levels at E(spl)-C (Figure 3F, S3C) while a CDK9 inhibitor, NVP2, had no effect in the same assays (Figure S3E-F). Third we investigated whether CDK module was recruited to E(spl)-C, in Notch-ON nuclei using an existing line in which YFP is fused to Med13 63,64. Significant enrichment of Med13::YFP was detected at E(spl)-C in Notch-ON nuclei, (Figure 3H) demonstrating that it is present in the hub with Mam.

Despite their effects on Mam recruitment, neither depletion of Med13 or CDK8 nor Senexin treatments caused any significant reduction in the levels of CSL recruited. On the contrary, a small increase occurred (Figure 3E, F). The differences in the effects on Mam and CSL imply that the CDK module is specifically involved in retaining Mam in the hub, and that in its absence other CSL complexes “win-out”, either because the altered conditions favour them and/or because they are the more abundant.

Mastermind is not essential for chromatin accessibility

The observed differences between CSL and Mam in their dynamics and dependency on CDK module led us to speculate that they make different functional contributions to the transcription hub. To probe the role of Mam, we examined the consequences when its recruitment to the complex was prevented. The N-terminal peptide from Mam functions as a dominant negative (MamDN) by occupying the groove formed by CSL-NICD and, when overexpressed, prevents transcription of target genes 8,9,65. As predicted, MamDN expression in Notch-ON conditions prevented recruitment of full-length Mam to E(spl)-C (Figure 4A). To verify that MamDN also inhibited transcription under these conditions, we used single molecule Fluorescent in situ hybridization (smFISH) with probes targeting E(spl)-C transcripts that are robustly up-regulated in this tissue 34. Nascent transcripts of E(spl)m3 were detected at E(spl)-C in Notch-ON nuclei and were at very reduced levels in nuclei co-expressing MamDN (Figure 4B) confirming that target-gene transcription is blocked.

Mam is not necessary for CSL enrichment and chromatin opening but necessary for transcription and Mediator 13 enrichment.

(A) Recruitment of GFP::Mam in Notch-ON nuclei is perturbed by MamDN (1151-Gal4 UAS-MamDN) (A) while recruitment of GFP::CSL is unaffected (A’). Control UAS were included in Notch-ON combinations, details are provided in Table S4. Scale bars represent 5um.

(B) Accessibility of Notch regulated enhancers adjacent to E(spl)bHLH-mβ and E(spl)bHLH-m3 under the conditions indicated was probed using ATAC-qPCR. Values were normalised to Rab11, error bars represent SEM. Closed control is a noncoding genomic region in chromosome 3 and open control’ Eip78, an ecdysone responsive region active in L3 larval stage.

(C) Average enrichment and max enrichment of Med13::YFP at E(spl)-C in Notch-ON conditions only, and in combination with MamDN. Quantifications as in Figure 1B, OFF, n=28, ON, n = 31. For p-values see Table S3

(D-D’) Expression of E(spl)bHLH-m3, detected by smFISH in the conditions indicated; representative images from cytoplasm (upper; 1.8mm3) and nucleus (lower; centred on E(spl)-C, blue arrows). Scale bars represent 5um. Graphs (D’), number of RNA puncta (cytoplasmic, n = 18, 20 and 22) and locus intensity (nucleus, n = 6, 7 and 6) in Notch OFF, Notch ON ctrl and Notch ON MamDN as indicated, boxplots as described in Figure 1B.

See also Supplementary Figure S4.

By contrast, recruitment of CSL was unaffected by MamDN expression. It was robustly recruited to E(spl)-C at a similar level to untreated Notch-ON nuclei despite the absence of full length Mam (Figure 4A’ 34). Identical results were obtained when Mam was depleted by RNAi (Figure S4A 34). The fact that CSL was still strongly enriched under these conditions argues that some of the Notch induced changes at E(spl)-C occur independently of the co-activator, as proposed previously 34. In contrast to CSL, the enrichment of Med13 was lost in the presence of MamDN revealing that Med13 recruitment requires Mam, as well as the converse (Figure 4C).

Since MamDN does not prevent CSL recruitment, although it blocks recruitment of Med13 and transcription, we hypothesised that some Notch-induced effects at target enhancers may not require Mam. One consequence from Notch activation is an increase in chromatin accessibility at sites where CSL is recruited 34,66. We therefore asked whether changes in chromatin accessibility still occur in the presence of MamDN by performing ATAC and analysing by qPCR the accessibility of the regulatory regions abutting E(spl)mβ and E(spl)m334. Both regions significantly increased in accessibility in Notch-ON nuclei which was maintained in the presence of MamDN (Figure 4D, D’). In addition, we found that the levels of GFP::NICD-IDR recruited to E(spl)-C were similar in the presence and absence of MamDN (Figure S4B), arguing that a modified hub is still formed.

Altogether these observations support the model that the increased chromatin accessibility elicited by Notch activation can occur independently of Mam and must rely on other functions conferred by CSL and NICD 34, while Mam is essential to recruit Mediator CDK module and enable transcription.

CSL recruitment and chromatin accessibility persist after Notch inactivation, conferring memory

Our results suggest that there are two or more separable steps involved in forming an active transcription hub in Notch-ON cells: a Mam-independent change in chromatin accessibility, a Mam-dependent recruitment of Mediator and initiation of transcription. If these are discrete steps, we reasoned that they might decay with different kinetics when Notch activity is removed. We took advantage of the thermosensitive Gal4/Gal80ts system to switch off Notch activity and assessed the consequences on CSL and Mam recruitment at two different time points: 4 hours and 8 hours after the switch-off. Imaging Mam::GFP and CSL::mCherry simultaneously, it was evident that Mam recruitment levels decreased more rapidly. CSL remained relatively constant through both time points while, in contrast, the levels of Mam at E(spl)-C decreased sharply after 4 hours (Figure 5A). Based on these results we propose that, after Notch activity decays, the locus remains accessible because when Mam-containing complexes are lost they are replaced by other CSL complexes (e.g. co-repressor complexes).

Effects of Notch withdrawal on hub composition imply memory state.

(A) Temperature paradigm: Withdrawal of Notch activity using the thermosensitive Gal4/Gal80ts system; after a 24h ON period, larvae were transferred to non-permissive temperatures for the indicated periods (top) and enrichment of mCherry::CSL and GFP::Mam at E(spl)-C imaged. Average enrichment, enrichment profiles and max enrichment quantified as in Figure 1B. 24 hrs ON, n = 28, 24 hrs ON, 4 hrs OFF, n = 21, 24 hrs ON, 8 hrs OFF, n = 19. p-values are summarized in Table S3.

(B) Accessibility of enhancers at E(spl)bHLH-mβ and E(spl)bHLH-m3, in Notch-ON and after Notch withdrawal (8 hours) was probed using ATAC-qPCR, as in Figure 4B.

(C) Expression of E(spl)bHLH-m3, detected by smFISH in the conditions indicated; representative images from cytoplasm (upper; 1.8 mm3) and nucleus (lower; centred on E(spl)-C, blue arrow). Representative images are shown, scale bars represent 5um. Graphs show number of RNA puncta (cytoplasmic) from n = 22, 30 and 30 regions, with boxplots as described before (Figure 1B). p-values in Table S3.

(D) Optogenetic paradigm: Conditions used to switch ON and OFF Notch activity using OptiC Notch 67.

(E) Recruitment of CSL::GFP and Mam::Halo at E(spl)-C, measured following blue light activation for 2 hours (ON) and after 4 hours in dark (OFF). GFP::CSL remains enriched, Halo::Mam is depleted.

(F) Recruitment of Halo::Mam in preactivated nuclei compared to naïve nuclei; temporal profile of Mam::Halo enrichment (right) and time taken to max enrichment (left). Error bars in the profile represents the SEM, values were normalised from 0 to 1 (similar to FRAP recovery, see Figure 1 and methods), n indicates the number of nuclei from >3 salivary glands. Boxplots parameters as in Figure 1B; naïve, n = 11, preactivated n = 9.

As Notch removal leads to a loss of Mam, but not CSL, from the hub, it should recapitulate the effects of MamDN. We therefore measured the accessibility of target enhancers at E(spl)mβ and E(spl)m3 at the 8-hour timepoint by ATAC. Neither enhancer exhibited any reduction in accessibility at this timepoint, consistent with the continued recruitment of CSL and the results obtained with MamDN (Figure 5B). In contrast, the gradual loss of Mam complexes was accompanied by reduced transcription, as detected by smFISH. Expression of E(spl)m3 was already significantly reduced at 4 hours, when levels of nascent transcripts at E(spl)-C and of cytoplasmic transcripts had. both decreased (Figure 5C). By 8 hours both nascent and cytoplasmic E(spl)m3 transcripts were almost undetectable (Figure 5C). Thus, the changes in transcription correlate well with the reduction in Mam recruitment whereas the chromatin accessibility persists in the absence of Mam.

We next tested whether target loci retain “memory” of previous Notch activation that would make them more receptive to a subsequent exposure to Notch activity. To do so, we took advantage of the temporal control provided by optogenetic release of NICD using OptIC-Notch{ω} 67. We compared the response to Notch activation in cells that had been ‘preactivated’ with blue light and subsequently kept in the dark for a short period, and ‘naïve’ cells, which had been kept solely in the dark and hence had no prior Notch activity (Figure 5D). After 4h in the dark following the initial activation, we observed that CSL was retained at E(spl)-C, while Mam was depleted, similar to the temporal experiments above. During the subsequent activation and imaging, enrichment of Mam occurred much more rapidly in preactivated cells in comparison to naïve cells, suggesting the former preserve a memory of previous activation and are primed to rapidly reform an active hub containing Mam (Figure 5F).

Together, our data indicate that CSL recruitment and increased chromatin accessibility persist after Notch removal and after the loss of the Mam activation complexes. This confers a memory state that enables rapid re-assembly of an activation hub in response to subsequent Notch activity.

Probability of transcription conferred by Mastermind

When analyzing the smFISH data we noticed that, even in Notch-ON conditions, a fraction of nuclei lacked foci of nascent transcription at E(spl)-C. Since Mam was present at E(spl)-C in all nuclei (Figure 6A, S5A), this led us to question whether the presence of Mam-complexes was sufficient to recruit downstream factors required for transcription initiation. We therefore investigated the extent that RNA Pol II was recruited to E(spl)-C in Notch-ON cells, using endogenous GFP::Rbp3 16. When scanning all nuclei, it was evident that Rbp3 enrichment at E(spl)-C in Notch-ON cells was highly variable. Robust enrichment was detected in a subset of nuclei but, in other cases, there was little/no enrichment above nuclear levels (Figure S5B). Performing a Gaussian population fitting on the data, 2 populations gave the best fit and, based on these, 36% of nuclei had significant enrichment and 64% had similar levels to Notch-OFF (Figure 6B, S5B). This differs from Mam where all nuclei fall into a single population that has significant Mam enrichment at E(spl)-C (Figure 6A, S5A). The striking difference in the proportions of nuclei with Pol II and with Mam enrichment implies that there is a limiting step, which results in transcription initiation being probabilistic/stochastic in these conditions.

Recruitment of Pol II and Med1 in Notch-ON is infrequent and is augmented by ecdysone.

(A) GFP::Mam enrichment at E(spl)-C in Notch-OFF and Notch-ON conditions as in Figure 1, represented with a violin plot with mean marked by a cross and median by a bar. Right graph: proportions of nuclei with significant enrichment (obtained by performing a Gaussian fit on the fluorescence intensity Images, see Supplementary Figure S5). Over 90% nuclei have significant enrichment (Notch-OFF, n = 32, Notch-ON, n = 30)

(B) Recruitment of GFP::Rbp1, a subunit of Pol II, at E(spl)-C in Notch-OFF and Notch-ON nuclei. Images: example where enrichment is detected. Graph: proportions of nuclei with significant enrichment as in A (Notch-OFF, n = 37, Notch-ON, n = 43). All scale bars represent 5um.

(C) Recruitment of GFP::Med1 at E(spl)-C in Notch-OFF and Notch-ON nuclei. Images: example where enrichment is detected. Graph: proportions of nuclei with significant enrichment as in A (Notch-OFF and ON, n = 38, 62). All scale bars represent 5um.

(D) Transcriptional activity in Notch-ON cells detected using live imaging of GFP::MCP recruited to MS2 loops in transcripts produced by E(spl)bHLHmβ (see methods). Images: representative examples of active (upper) and inactive (lower) nuclei. Graph: Proportion of active nuclei (Notch-ON, n = 53)

(E) Recruitment of GFP::Med1, mCherry::Rbp1 and Halo::CSL. Med1 and Rbp1 recruitments are correlated. Images: example of co-recruitment of Med1, Rbp1 and CSL. Graph: Correlation plot of max enrichment per nucleus at E(spl)-C, (marked by Halo::CSL recruitment), n = 48.

(F) Pretreatment with ecdysone in Notch-ON conditions increases proportions of nuclei with recruitment of GFP::Rbp1, GFP::Med1 and with active transcription foci (MS2/MCP intensity) compared to controls (ETOH). Proportions of nuclei with significant enrichment defined as in A. Rbp3 ctrl, ecdysone: n = 39, 36. Med1 ctrl, ecdysone: n = 34, 34. MS2 ON ctrl, ON ecdysone, OFF ecdysone: n = 52, 42, 22.

See also Supplementary Figure S5.

It has been suggested that the core Mediator complex, as distinct from the CDK module, has an important role in bridging between enhancer bound transcription complexes and initiation complexes at promoters 68–70. To investigate core Mediator recruitment we generated a GFP::Med1 fusion by CRISPR genome-editing. Using this endogenously tagged protein, which is homozygous viable, we found that GFP::Med1 became enriched at E(spl)-C with a similar probability to Rbp3 (Figure 6C). Indeed when we compared directly the enrichment of Pol II and Med1 in Notch-ON nuclei, by using mCherry::Rbp1 16 to monitor Pol II, we observed a significant correlation between the two proteins (R2 = 0.69) (Fig 6E). No such correlation with CSL was observed, in the same experiments (Figures S5E).

To verify that only a subset of nuclei was transcriptionally active we used a strain where MS2 loops have been inserted into E(spl)mβ using CRISPR-Cas9 engineering 11 and imaged transcription live in Notch-ON nuclei. In the presence of MCP-GFP, which binds to MS2 loops in the RNAs produced, nascent sites of transcription appear as puncta that align into a band of fluorescence at E(spl)-C, due to the multiple aligned gene copies. A robust band of MCP/MS2 nascent transcription at E(spl)-C was detected in Notch-ON conditions, demonstrating that some nuclei were actively transcribing (Figure 6D, S5C). However, this was present in only a third (37%) of nuclei. These data show that only a subset of nuclei engage in active transcription, and that the proportion of active nuclei is similar to that with recruitment of Pol II and Med1.

These data demonstrate that presence of Mam-complexes is not sufficient to reliably drive all the steps required for transcription in every Notch-ON nucleus. Instead, it appears that transcription is initiated stochastically. This is unexpected and could only be detected because our in vivo live imaging approaches enable us to monitor the responses of individual nuclei.

Ecdysone cooperates with Notch to increase probability of transcription

Even though relatively few Notch-ON nuclei became transcriptionally active in our experiments, they all had robust recruitment of Mam complexes and of Med13 at E(spl)-C. Thus, in many respects the gene locus becomes competent or poised for transcription in all nuclei. We wondered, therefore, whether the presence of a second stimulatory signal would increase the probability of loci becoming transcriptionally active. In normal development, salivary glands become exposed to the steroid hormone ecdysone a few hours after we perform our experiments. Two observations suggested that ecdysone was a good candidate for a cooperating signal. First, previous genome-wide studies detected ecdysone receptor binding in the E(spl)-C region 71. Second, we noticed in rare samples containing an older gland, that it had a higher proportion of active nuclei. We therefore exposed the early-stage Notch-ON salivary glands to ecdysone and analyzed the proportion of nuclei with MS2/MCP puncta. Strikingly, the proportion of active nuclei increased dramatically to 70% following ecdysone treatment. No such effects were seen in Notch-OFF nuclei where E(spl)mβ remained silent even after ecdysone treatment (Figure 6F).

We further tested the effect of ecdysone by measuring changes in enrichment of Pol II and Med1 at E(spl)-C. The proportion of nuclei with clear enrichment of each factor was significantly increased (Figure 6F). Thus, the switch from the more stochastic transcription elicited by recruitment of Mam alone to the more robust initiation when ecdysone was added, correlated with the presence of Med1 and Pol II. As Mediator is reported to stabilize enhancer-promoter interactions, its recruitment may be what limits the probability of transcription 68,72.

Discussion

Understanding the mechanisms by which signaling levels and dynamics are accurately converted to transcriptional outputs is fundamental for developmental programming. Here we used live imaging approaches to probe how this occurs in the context of the Notch pathway, where the transcriptional relay relies on the single NICD released by each activated receptor and the nuclear complexes it forms with its partners CSL and Mam. By tracking these complexes in real time we unveiled three important features (Figure 7): (i) in Notch-ON nuclei, the activation complexes promote formation of a dynamic protein hub at a regulated gene locus that concentrates key factors including Mediator CDK module, (ii) the composition of the hub is changeable and the footprint persists after Notch withdrawal conferring a memory that enables rapid reactivation, (iii) transcription is probabilistic in Notch-ON nuclei, such that only a third of nuclei with a hub are actively transcribing. This has far-reaching implications because it reveals that stochastic differences in Notch pathway output can arise downstream of receptor activation.

Model illustrating different modes of Notch transcription-hub.

In the absence of Notch activity (Notch-OFF) target genes are inactive, CSL is complexed with co-repressors (e.g. Hairless). Following Notch activation (Notch ON), released NICD (purple) generates a localized high concentration of transcription factors around target enhancer(s) referred to as a “hub” (orange). Open hub: CSL (green) recruitment and accessible chromatin can occur in the absence of Mam. Engaged hub: presence of Mam (magenta) favours recruitment of additional factors, including Mediator CDK8 module (yellow). Active hub: productive transcription, transition to this mode, with core Mediator (yellow) and Pol II (pale blue) enrichment, is stochastic (dotted arrow). The probability can be enhanced by secondary signal, such as provided by ecdysone. Memory hub: CSL enrichment and chromatin accessibility remain after withdrawal of NICD.

Notch transcription complexes form a hub

Many recent studies have demonstrated that hubs or condensates play key roles in gene-expression regulation by maintaining a high local concentration of transcription factors and other regulatory complexes 17,22,47,68. By tracking the co-activator Mam in real time, we show that Notch transcription complexes become enriched around a target locus in a zone that has characteristics of a hub. First, the high-concentration zones are highly dynamic and undergo a continual exchange between activator and repressor complexes. The former exhibit longer residence times, suggesting they are stabilised by other interactions. Second, recruitment of CSL-containing complexes is non-stoichiometric with respect to CSL-binding motifs and can in part be recapitulated by an IDR from NICD. Although deletion of the IDR biases against activation complex recruitment, our evidence suggests that individual IDRs make a minor contribution, as observed recently in an unbiased study of IDRs. Third, the enriched zone is readily detectable by live imaging, but is sensitive to fixation, as has been reported for some subcellular protein condensates 73. Although transcription hubs have frequently been associated with phase separation, examples are emerging where high local concentrations of transcription factors can be achieved without it 21. Similarly, the locally enriched Notch-induced hubs in our experiments do not manifest clear properties of liquid condensates despite being non-stoichiometric.

We propose that NICD and Mam participate in multivalent interactions which increase the likelihood of the tripartite activation complexes remaining in the chromatin-associated state and which facilitate recruitment of coactivators to form a local hub (Figure 7). The latter include subunits of CDK module, Med 13. We find that perturbations to CDK module compromise Mam levels in the hub and vice versa, highlighting its importance. Indeed, the CDK module has been reported to interact directly with Mam and to play a role in Notch signalling output 31,32, and recent studies in mammalian cells indicate CDK8 acts positively in signal-induced gene expression 74. Although its recruitment is proposed to prepare genes for activation, the transition to activation necessitates CDK8 release, since CDK module sterically inhibits core Mediator-Pol II interactions72,75. This is consistent with our results showing that Med13 recruitment is highly dynamic and that, despite most nuclei showing Med13 enrichment within the hub, the probability of transcription is low, as discussed below.

Role of hub in conferring transcriptional memory

Assembly of the transcription hub at E(spl)-C depends on Notch activity but, surprisingly, can form in the absence of the co-activator Mam, albeit the composition differs. Notably, CSL complexes continue to be highly enriched and the chromatin at E(spl)-C enhancers is accessible in the absence of Mam (Figure 7). These properties also persist for several hours after withdrawal of Notch activity and in this way, enhancers that have been switched on by Notch retain an imprint that can influence their response to a subsequent signal, a phenomenon known as transcriptional memory76,77. As an indication, after a previous Notch activation event, the conversion to an activation hub with Mam enrichment occurs more rapidly than in the absence of a prior active state. Similar accelerated recruitment of STAT1 to promoters occurred in cells with IFNγ-induced transcriptional memory 78, and augmented transcription levels have been seen in several contexts 79,80.

The concept that enhancers retain a memory of a recent Notch signal has several implications. First, naïve and pre-activated loci will respond with different kinetics, which could bias cell-fate decisions when iterative Notch signals are deployed 81. Second, although our analysis has focussed on relatively short time scales (up to 8 hours), the footprint may be inherited by daughter cells as CSL is retained on mitotic chromosomes in at least some cell-types 82. If this is the case, it may explain why enhancer decommissioning is important to switch-off the Notch response in some contexts 83–85. Third, it could explain some of the observed recruitment of CSL corepressor complexes to active chromatin if they are involved in sustaining the hub when the signal decays 86,87.

Transcription initiation is probabilistic

Because the in vivo live imaging enables us to monitor the responses of individual nuclei, we made the surprising discovery that the presence of Mam-containing hubs at E(spl)-C is not sufficient to guarantee transcription. Instead, our results reveal that Mam recruitment confers a 1 in 3 chance of transcription. At any one time, only a third of nuclei are actively transcribing. This has profound implications because it has widely been assumed that stochastic differences in Notch pathway output arise due to fluctuations in ligands or ligand availability 88,89. Our results raise the possibility that differences may also arise due to intrinsic variations downstream of receptor activation that affect the probability of transcription occurring. Furthermore, this probability can be modified by other signals, as we observed here with ecdysone, or by the presence of other transcription factors, as when enhancers are ‘primed’ 90 and/or cooperatively regulated 91–95.

In our experiments, the probability of transcription correlated with the proportion of nuclei where Med1, a core Mediator subunit, was recruited into the hub at E(spl)-C. Mediator is an important intermediary between transcription factors and the preinitiation complex and it functions as an ‘integrator’ coordinating diverse and combinatorial inputs 68. The increased probability of Mediator recruitment can thus be explained by an increase in the valency of possible interactions when Notch and ecdysone are both active and present at the enhancers of target genes. We propose that this will be a general mechanism and that other signals/transcription factors will synergise with Notch complexes by adding to the valency of interactions and facilitating recruitment of coactivators to increase the probability of transcription. In different contexts this could toggle the response from one that is stochastic to one that is hard-wired.

Supporting information

eLife Assessment

This fundamental study advances our understanding of how Notch signaling activates transcription by analyzing the dynamics of the Mastermind transcriptional co-activator and its role in the activation complex. The evidence is compelling with precise quantitative measurements. The evidence presented by the authors features methods, data, and analyses that are currently state-of-the-art but have not previously been applied to how transcription is regulated in the Notch pathway.

Reviewer #1 (Public Review):

In this manuscript by DeHaro-Arbona et al., the authors wish to understand how a signaling pathway (Notch) is dynamically decoded to elicit a specific transcriptional output. In particular, they investigate the kinetic properties of Notch-responsive nuclear complexes (the DNA binding factor CSL and its co-activator Mastermind (mam) along with several candidate interacting partners). Their experimental model is the polytene chromosome of the Drosophila salivary gland, in which the naturally inactive Notch can be artificially induced through the expression of a constitutively active form of Notch.

The authors develop a series of CRISPR and transgenic lines enabling the live imaging of these complexes at a specific locus and in various backgrounds (genetic perturbations/drug treatments). This quantitative live imaging data suggests that Notch nuclear complexes form hubs and the authors characterize their binding dynamics. Interestingly, they elegantly demonstrate that the content of these hubs and their kinetic properties can evolve, even within Notch ON cells. Hence, they propose the existence of distinct hubs, distinguishing an open (CSL), engaged (CSK-Mam), or active (CSL-Mam-Med-PolII) configuration in Notch ON cells and an inactive hub (in Notch OFF having previously been exposed to Notch) state, that would explain the surprising transcriptional memory that the authors observe hours after Notch withdrawal.

Reviewer #2 (Public Review):

The manuscript from deHaro-Arbona et al, entitled "Dynamic modes of Notch transcription hubs conferring memory and stochastic activation revealed by live imaging the co-activator Mastermind", uses single molecule microscopy imaging in live tissues to understand the dynamics and molecular determinants of transcription factor recruitment to the E(spl)-C locus in Drosophila salivary gland cells under Notch-ON and -OFF conditions. Previous studies have identified the major players that are involved in transcription regulation in the Notch pathway, as well as the importance of general transcriptional coregulators, such as CBP/P300 and the Mediator CDK module, but the detailed steps and dynamics involved in these processes are poorly defined. The authors present a wealth of single molecule data that provides significant insights into Notch pathway activation, including:

1. Activation complexes, containing CSL and Mam, have slower dynamics than the repressor complexes, containing CSL and Hairless.

2. Contribution of CSL, NICD, and Mam IDRs to recruitment.

3. CSL-Mam slow-diffusing complexes are recruited and form a hub of high protein concentrations around the target locus in Notch-ON conditions.

4. Mam recruitment is not dependent on transcription initiation or RNA production.

5. CBP/P300 or its associated HAT activity is not required for Mam recruitment.

6. Mediator CDK module and CDK8 activity are required for Mam recruitment, and vice-versa, but not CSL recruitment.

7. Mam is not required for chromatin accessibility but is dependent on CSL and NICD.

8. CSL recruitment and increased chromatin accessibility persist after NICD removal and loss of Mam, which confers a memory state that enables rapid re-activation in response to subsequent Notch activation.

9. Differences in the proportions of nuclei with both Pol II and with Mam enrichment, which results in transcription being probabilistic/stochastic. These data demonstrate that the presence of Mam-complexes is not sufficient to drive all the steps required for transcription in every Notch-ON nucleus.

10. The switch from more stochastic to robust transcription initiation was elicited when ecdysone was added.

Overall, the manuscript is well written, concise, and clear, and makes significant contributions to the Notch field, which are also important for a general understanding of transcription factor regulation and behavior in the nucleus. I recommend that the authors address my relatively minor criticisms detailed below.

Page 7, bottom. The authors speculate, "It is possible therefore that, once recruited, Mam can be retained at target loci independently of CSL by interactions with other factors so that it resides for longer." Is it possible that another interpretation of that data is that Mam is a limiting factor?

Page 9. The authors write, "A very low level of enrichment was evident for... for the CSL C-terminus..". The recruitment of CSL ct IDR does not appear to be statistically significant or there is no apparent difference (Figure S2C), suggesting the CSL ct IDR does not play a role in enrichment.

Page 9. The authors write, "Notably, MamnIDR::GFP fusion was present in droplets, suggesting it can self-associate when present in a high local concentration (Figure S2B)." Is this result only valid for Mam nIDR or does full-length Mam also localize into droplets, as has been previously observed for full-length mammalian Maml1 in transfected cells?

Previous studies in mammalian cells suggest that Maml1 is a high-confidence target for phosphorylation by CDK8, see Poss et al 2016 Cell Reports https://doi.org/10.1016/j.celrep.2016.03.030. By sequence comparison, does fly Mam have similar potential phosphorylation sites, and might these be critical for Mam/CDK module recruitment?

Page 11: The authors write, "The differences in the effects on Mam and CSL imply that the CDK module is specifically involved in retaining Mam in the hub, and that in its absence other CSL complexes "win-out", either because the altered conditions favour them and/or because they are the more abundant." Are the "other" complexes the authors are referring to Hairless-containing complexes? With the reagents the authors have in hand couldn't this be explicitly shown for CSL-complexes rather than speculated upon?

Page 12/13: The authors write, "Based on these results we propose that, after Notch activity decays, the locus remains accessible because when Mam-containing complexes are lost they are replaced by other CSL complexes (e.g. co-repressor complexes)." Again, why not actually test this hypothesis rather than speculate? The dynamics of Hairless complexes following the removal of Notch would be very interesting and build upon previously published results from the Bray lab.

Page 13: The authors write, "As Notch removal leads to a loss of Mam, but not CSL, from the hub, it should recapitulate the effects of MamDN." While the data in Figure 5B seem to support this hypothesis, it's not clear to me that the loss of Mam and MamDN should phenocopy each other, bc in the case of MamDN, NICD would still be present.

The temporal dynamics for Mam recruitment using the temperature- and optogenetic-paradigms are quite different. For example, in the optogenetic time course experiments, the preactivated cells are in the dark for 4 hours, while in the temperature-controlled experiments, there is still considerable enrichment of Mam at 4 hours. For the preactivated optogenetic experiments, how sure are the authors that Mam is completely gone from the locus, and alternatively, can the optogenetic experimental results be replicated in the temperature-controlled assays? My concern is whether the putative "memory" observation is just due to incomplete Mam removal from the previous activation event.

Reviewer #3 (Public Review):

Summary:

DeHaro-Arbona and colleagues investigate the in vivo dynamics of Notch-dependent transcriptional activation with a focus on the role of the Mastermind (MAM) transcriptional co-activator. They use GFP and HALO-tagged versions of the CSL DNA-binding protein and MAM to visualize the complex, and Int/ParB to visualize the site of Notch-dependent E(Spl)-C transcription. They make several conclusions. First, MAM accumulates at E(Spl)-C when Notch signaling is active, just like CSL. Second, MAM recruits the CDK module of Mediator but does not initiate chromatin accessibility. Third, after signaling is turned off, MAM leaves the site quickly but CSL and chromatin accessibility are retained. Fourth, RNA pol II recruitment, Mediator recruitment, and active transcription were similar and stochastic. Fifth, ecdysone enhances the probability of transcriptional initiation.

Strengths:

The conclusions are well supported by multiple lines of extensive data that are carefully executed and controlled. A major strength is the strategic combination of Drosophila genetics, imaging, and quantitative analyses to conduct compelling and easily interpretable experiments. A second major strength is the focus on MAM to gain insights into the dynamics of transcriptional activation specifically.

Weaknesses:

Weaknesses are minor. There were no p-values reported for data presented in Figure S1D and no indication of how variable measurements were. In addition, the discussion of stochasticity was not integrated optimally with relevant literature.

Author Response

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

In this manuscript by DeHaro-Arbona et al., the authors wish to understand how a signaling pathway (Notch) is dynamically decoded to elicit a specific transcriptional output. In particular, they investigate the kinetic properties of Notch-responsive nuclear complexes (the DNA binding factor CSL and its co-activator Mastermind (mam) along with several candidate interacting partners). Their experimental model is the polytene chromosome of the Drosophila salivary gland, in which the naturally inactive Notch can be artificially induced through the expression of a constitutively active form of Notch.

The authors develop a series of CRISPR and transgenic lines enabling the live imaging of these complexes at a specific locus and in various backgrounds (genetic perturbations/drug treatments). This quantitative live imaging data suggests that Notch nuclear complexes form hubs, and the authors characterize their binding dynamics. Interestingly, they elegantly demonstrate that the content of these hubs and their kinetic properties can evolve, even within Notch ON cells. Hence, they propose the existence of distinct hubs, distinguishing an open (CSL), engaged (CSK-Mam), or active (CSL-Mam-Med-PolII) configuration in Notch ON cells and an inactive hub (in Notch OFF having previously been exposed to Notch) state, that would explain the surprising transcriptional memory that the authors observe hours after Notch withdrawal.

We thank the reviewer for this constructive summary of our work

Reviewer #2 (Public Review):

The manuscript from deHaro-Arbona et al, entitled "Dynamic modes of Notch transcription hubs conferring memory and stochastic activation revealed by live imaging the co-activator Mastermind", uses single molecule microscopy imaging in live tissues to understand the dynamics and molecular determinants of transcription factor recruitment to the E(spl)-C locus in Drosophila salivary gland cells under Notch-ON and -OFF conditions. Previous studies have identified the major players that are involved in transcription regulation in the Notch pathway, as well as the importance of general transcriptional coregulators, such as CBP/P300 and the Mediator CDK module, but the detailed steps and dynamics involved in these processes are poorly defined. The authors present a wealth of single molecule data that provides significant insights into Notch pathway activation, including:

(1) Activation complexes, containing CSL and Mam, have slower dynamics than the repressor complexes, containing CSL and Hairless.

(2) Contribution of CSL, NICD, and Mam IDRs to recruitment.

(3) CSL-Mam slow-diffusing complexes are recruited and form a hub of high protein concentrations around the target locus in Notch-ON conditions.

(4) Mam recruitment is not dependent on transcription initiation or RNA production.

(5) CBP/P300 or its associated HAT activity is not required for Mam recruitment.

(6) Mediator CDK module and CDK8 activity are required for Mam recruitment, and vice-versa, but not CSL recruitment.

(7) Mam is not required for chromatin accessibility but is dependent on CSL and NICD.

(8) CSL recruitment and increased chromatin accessibility persist after NICD removal and loss of Mam, which confers a memory state that enables rapid re-activation in response to subsequent Notch activation.

(9) Differences in the proportions of nuclei with both Pol II and with Mam enrichment, which results in transcription being probabilistic/stochastic. These data demonstrate that the presence of Mamcomplexes is not sufficient to drive all the steps required for transcription in every Notch-ON nucleus.

(10) The switch from more stochastic to robust transcription initiation was elicited when ecdysone was added.

Overall, the manuscript is well written, concise, and clear, and makes significant contributions to the Notch field, which are also important for a general understanding of transcription factor regulation and behavior in the nucleus. I recommend that the authors address my relatively minor criticisms detailed below.

We thank the reviewer for their thorough and constructive summary of our work. We are glad that they overall found it insightful and interesting. Below we have addressed the points they have raised.

Page 7, bottom. The authors speculate, "It is possible therefore that, once recruited, Mam can be retained at target loci independently of CSL by interactions with other factors so that it resides for longer." Is it possible that another interpretation of that data is that Mam is a limiting factor?

As indicated our comment is a speculation and is based on the observations summarized in the paragraph. We are not entirely sure what the reviewer is proposing as an alternate model. However, if it relates to the relative concentrations of the different factors, this would not account for the differences in trajectory durations. And for most aspects of our analysis, K[off] has the most profound influence on the results. Furthermore, differences persist even when CSL levels are considerably reduced (as in conditions with Hairless RNAi).

Page 9. The authors write, "A very low level of enrichment was evident for... for the CSL Cterminus..". The recruitment of CSL ct IDR does not appear to be statistically significant or there is no apparent difference (Figure S2C), suggesting the CSL ct IDR does not play a role in enrichment.

We agree with the comments of the reviewer and have adjusted the text on page 9 accordingly.

Page 9. The authors write, "Notably, MamnIDR::GFP fusion was present in droplets, suggesting it can self-associate when present in a high local concentration (Figure S2B)." Is this result only valid for Mam nIDR or does full-length Mam also localize into droplets, as has been previously observed for full-length mammalian Maml1 in transfected cells?

We agree that the observed foci of MamL1 that have been detected in mammalian cells are interesting. We have not tried to replicate those data because the large size of Mam has made it challenging to produce a full-length form in over-expression. We note however that another portion of Mam, MamIDR, does not make droplets when over-expressed despite it containing a large section of the disordered region of the Drosophila Mam. We have now included a comment about the mammalian data in the text (page 9) to put our findings in context.

Previous studies in mammalian cells suggest that Maml1 is a high-confidence target for phosphorylation by CDK8, see Poss et al 2016 Cell Reports https://doi.org/10.1016/j.celrep.2016.03.030. By sequence comparison, does fly Mam have similar potential phosphorylation sites, and might these be critical for Mam/CDK module recruitment?

We thank the reviewer for highlighting this point. Indeed, we were very excited when we learnt that MamL1 was found to be a high confidence CDK8 target and we looked hard in the Mam sequence for potential phosphorylation sites. Sadly, there is very little conservation between the fly and the mammalian proteins beyond the helical region that contacts CSL and NICD. Furthermore, there are no identifiable putative CDK8 phosphorylation sites based on conventional motifs. It therefore remains to be established whether or not Mam is a direct target of the CDK8 kinase activity. We have added an explanatory comment in the text (page 11).

Page 11: The authors write, "The differences in the effects on Mam and CSL imply that the CDK module is specifically involved in retaining Mam in the hub, and that in its absence other CSL complexes "win-out", either because the altered conditions favour them and/or because they are the more abundant." Are the "other" complexes the authors are referring to Hairless-containing complexes? With the reagents the authors have in hand couldn't this be explicitly shown for CSLcomplexes rather than speculated upon?

The reviewer is correct that CSL complexes containing Hairless are good candidates to be recruited in these conditions. We have compared the levels of Hairless at E(spl)-C following treatments with Senexin and have not detected a difference. However, it appears that the high proportion of unbound Hairless makes it difficult to detect/quantify the enrichment at E(spl)-C. We have therefore taken a different strategy, which is to measure the recruitment of a mutant form of CSL that is compromised for Hairless binding. Recruitment of the mutant CSL is detected in Notch-ON conditions, but is significantly reduced/absent following Senexin treatment. These data favour the model proposed by the reviewer that in the absence of CDK8 activity, the CSL-Hairless complexes win out. These new data have been added in new Supplementary Figure S3F and S3G (and see text page 11)

Page 12/13: The authors write, "Based on these results we propose that, after Notch activity decays, the locus remains accessible because when Mam-containing complexes are lost they are replaced by other CSL complexes (e.g. co-repressor complexes)." Again, why not actually test this hypothesis rather than speculate? The dynamics of Hairless complexes following the removal of Notch would be very interesting and build upon previously published results from the Bray lab.

We thank the reviewer for this comment and we agree it’s possible that the proportion of Hairless complexes increases after Notch withdrawal. However, for the reasons outlined above, it is difficult to quantify changes in Hairless, (and our preliminary experiment did not reveal any large-scale effect) and because of the complexity of the genetics we cannot straightforwardly extend the experiment to analyze the behaviour of the mutant CSL as above. Therefore, at present, we cannot say whether the loss of Mam is compensated by an increase in Hairless. We hope in future to investigate the characteristics of the memory in more depth.

Page 13: The authors write, "As Notch removal leads to a loss of Mam, but not CSL, from the hub, it should recapitulate the effects of MamDN." While the data in Figure 5B seem to support this hypothesis, it's not clear to me that the loss of Mam and MamDN should phenocopy each other, bc in the case of MamDN, NICD would still be present.

We apologise that this sentence was a bit misleading. We have now rewritten it to improve accuracy (page 13) “As Notch removal leads to a loss of Mam, but not CSL, from the hub, we hypothesised it would recapitulate the effects of MamDN on chromatin accessibility and transcription of targets.”

The temporal dynamics for Mam recruitment using the temperature- and optogenetic-paradigms are quite different. For example, in the optogenetic time course experiments, the preactivated cells are in the dark for 4 hours, while in the temperature-controlled experiments, there is still considerable enrichment of Mam at 4 hours. For the preactivated optogenetic experiments, how sure are the authors that Mam is completely gone from the locus, and alternatively, can the optogenetic experimental results be replicated in the temperature-controlled assays? My concern is whether the putative "memory" observation is just due to incomplete Mam removal from the previous activation event.

We appreciate the concerns of the reviewer. However, we are confident that the 4-hour optogenetic inactivation is much more effective than the equivalent time for temperature shifts. The temperature sensitive experiment involves a longer decay, because not only the protein but also the mRNA has to decay to fully remove NICD activity. The optogenetic experiments, involve only protein decay and so are more acute. Furthermore, we have tested (and we show in Figure 5H) that Mam is fully depleted after 4 hours “Off” in the optogenetic experiments.

In order to further strengthen the evidence in favour of the memory hub, we have extended the time-frame further to show that CSL is retained at the locus even after 24 hours “Notch OFF” in both the temperature and the optogenetic paradigm. We have also measured the effects on transcription after a 24hr OFF period using the optogenetic paradigm and seen that robust transcription is initiated in cells that have experienced a previous activation (preactivated) compared to those that have not (naïve). These new data have been added to new Figure 5 C-F and strongly support the memory model.

Reviewer #3 (Public Review):

Summary:

DeHaro-Arbona and colleagues investigate the in vivo dynamics of Notch-dependent transcriptional activation with a focus on the role of the Mastermind (MAM) transcriptional co-activator. They use GFP and HALO-tagged versions of the CSL DNA-binding protein and MAM to visualize the complex, and Int/ParB to visualize the site of Notch-dependent E(Spl)-C transcription. They make several conclusions. First, MAM accumulates at E(Spl)-C when Notch signaling is active, just like CSL. Second, MAM recruits the CDK module of Mediator but does not initiate chromatin accessibility. Third, after signaling is turned off, MAM leaves the site quickly but CSL and chromatin accessibility are retained. Fourth, RNA pol II recruitment, Mediator recruitment, and active transcription were similar and stochastic. Fifth, ecdysone enhances the probability of transcriptional initiation.

Strengths:

The conclusions are well supported by multiple lines of extensive data that are carefully executed and controlled. A major strength is the strategic combination of Drosophila genetics, imaging, and quantitative analyses to conduct compelling and easily interpretable experiments. A second major strength is the focus on MAM to gain insights into the dynamics of transcriptional activation specifically.

We thank the reviewer for their positive comments about the strengths of our work.

Weaknesses:

Weaknesses are minor. There were no p-values reported for data presented in Figure S1D and no indication of how variable measurements were. In addition, the discussion of stochasticity was not integrated optimally with relevant literature.

We thank the reviewer for noting these points. The statistical tests have now been included for Figure S1D (now Figure S1F). We have amplified the discussion about stochasticity, to include more reference to the literature and to make clear also the distinction with transcription bursting (page 19, 20).

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

The authors have an elegant series of manipulations that provide strong evidence for their hypotheses and conclusions. Their exploitation of a unique biological system amenable to imaging in the larval salivary gland is well-considered and well-performed. Most of the conclusions are supported by the data. I only have the concerns below.

(1) One of the main findings is the composition of Notch nuclear complexes and their interactions within a 'hub'. Yet most of the data showing hubs focus on labeling one protein component (+the locus or transcription), but multi-color imaging is rarely used to show how CSL-Mam, Mam-Med... protein signals coalescence to form a hub. Given the powerful tool developed, it would be important to show these multi-state hubs. Related to this, if the authors expect that hubs are formed independently of transcription or Notch pathway activation, do the authors see clustering at other non-specific loci in the nucleus? If not, can the authors comment on why they think that is the case? If so, do they demonstrate consistent residence time profiles with the tracked E(spl) locus?

We apologise that it was not evident from the data shown that the proteins co-localize. First we stress that all the experiments are multicolor and most rely on very powerful methods to measure co-recruitment at a chromosomal locus- something that is very rarely achieved by others studying hubs. Second, we have in all cases confirmed that the proteins do colocalize. We have modified the diagram of our analysis pipeline to make more clear that this relies on multi-colour imaging, and adjusted all the figure labels to indicate the position of E(spl)-C. We have also added panels to new supplementary Figure S1C with examples of the co-localization between CSL and Mam and a plot confirming their levels of recruitment are correlated across multiple nuclei.

We would like to clarify that our data show that the hubs do require Notch activation for their establishment. Other regions of enrichment are detected in Notch-ON conditions, but these are less prominent and, with no independent method for identifying them, can’t be compared between nuclei. In SPT experiments, other clusters with consistent residence are detected as reported in our recent paper which expanded on the SPT data (Baloul et al, 2023). We also detect co-localizations and “hubs” in other tissues, but those analyses are ongoing and beyond the scope of this paper.

(2) The authors convincingly show that Notch hub complexes exhibit a memory. While the data showing rapid hub reformation upon Notch withdrawal are solid and convincing (Figure 5, in particular, F), the claim that this memory fosters rapid transcriptional reactivation is less clear. Yet in order to invoke transcriptional memory, it's necessary to solidify this transcriptional response angle. The authors should consider quantifying the changes in transcription activity (at the TS and not in the cytoplasm as currently shown), as well as the timing of transcriptional reactivation (with the MS2 system or smFISH). Manipulating the duration of the activation and dark recovery periods could help to draw a better correlation between the timing of hub reformation and that of transcriptional response and would also help determine how persistent this phenomenon is.

We thank the reviewer for these suggestions. We have carried out several new experiments to probe further the persistence of memory and to show the effects on transcription when Notch is inactivated/reactivated. First, we have extended the time period for Notch inactivation by temperature control and show that the CSL hub persists even at 24 hours and that no transcription from the target E(spl)m3 is detected –neither at the transcription start-site nor in the cytoplasm. Second, we have extended the Notch OFF time period to 24 hours using the optogenetic approach and show that transcription is robustly reinitiated in preactivated nuclei when Notch is re-activated with 30 mins light treatment while little if any E(spl)m3 transcription is detected in naïve nuclei with the same treatment. These new data are included in new Figure 5 C-F and see page 13-14. Both these new experiments substantiate the model that the nuclei retain transcriptional memory.

(3) The manuscript ends with the finding that the presence of a Mam hub does not always correlate with transcription. They conclude that transcription is initially stochastic. The authors find this surprising and even state that this could not be observed without their in vivo live imaging approaches. I don't understand why this result is surprising or unexpected, as we now know that transcription is generally a stochastic process and that most (if not all) loci are transcribed in a bursting manner. The fact that E(spl)-C locus is bursty is already obvious from the smFISH data. The fact that active nascent transcription does not correlate with local TF hubs was already observed in early Drosophila embryos (with Zelda hubs and two MS2 reporters, hb-MS2, sna-MS2). If, in spite of the inherent stochasticity of transcription (bursting), the data are surprising for other reasons, the authors should explain it better.

We apologise that we had not made clear the reasons why the results were unexpected. We have substantially rewritten this section, and the discussion section, to clarify. We have also moderated the language used to better reflect the overall context of our results. We briefly summarise here. As the reviewer correctly states, it is well known that transcription is inherently bursty. Indeed the MS2 transcription profiles in “ON” nuclei are bursty, which likely reflects the switching of the promoter. However, in other contexts where we have monitored transcription although it is bursty it has nevertheless been initiated synchronously in response to Notch in all nuclei in a manner that was fully penetrant. What we observe in our current conditions, is that some nuclei never initiate transcription over the time-course of our experiments (2-3 hours), and those that are ON rarely switch off. This implies that there is another rate-limiting step. Supplying a second signal can modulate this so that it occurs with much higher frequency/penetrance. We consider this to be a second tier of regulation above the fundamental transcriptional bursting.

The fact that Mam is recruited in all nuclei, whether or not they are actively transcribing was surprising because recruitment of the activation complex has been considered as the limiting step. This is somewhat different from Zelda, which is thought to be permissive and needed at an early step to prime genes for later activation rather than to be the last step needed to fire transcription. We note also that we are not monitoring the position of the hub with respect to the promoter, as in the Zelda experiments (Zelda hubs may still persist, but they are not overlapping with the nascent RNA), we are monitoring the presence or absence of Mam hub in proximity to a genomic region.

Minor suggestions:

(1) The genotypes of the samples should be indicated in the figure legends.

We thank the reviewer for this suggestion. We have provided a table (new Table S3) where all of the genetic combinations are provided in detail for each figure. We considered that this approach would be preferable because it would be quite cumbersome to have the genotypes in each legend as they would become very long and repetitive.

(2) While the schematic Fig1A explains how the locus is detected, the presence of ParS/ParB is never indicated in subsequent panels and Figure. I assume that all panels depicting enrichment profiles, use a given radius from the ParS/ParB dot to determine the zero of the x-axis (grey zone). This should be clearly stated in all panels/figure legends concerned.

We apologies if this was not made explicit. Yes, all panels depicting enrichment profiles, use immunofluorescence signal from ParA/ParB recruitment to determine the zero of the x-axis. We have now marked this more clearly In all figures (grey bar, grey shading or labelled 0). All images where the locus is indicated by an arrowhead, by a coloured bar above the intensity plots or by grey shading in the graphs have been captured with dual colour and the signal from ParA/B recruitment used to define its location. This is now clearly stated in the analysis methods and in the legend. We have also modified the diagram in new supplementary Figure S1B, showing our analysis pipeline, to make that more explicit.

(3) FRAP/SPT experiments: the author should provide more details. How many traces? Are traces showing bleaching removed?

P7: does the statement ' The residences are likely an underestimation because bleaching and other technical limitations also affect track durations' imply that traces showing bleaching have not been removed from the analysis?

The authors could justify the choice of the model for fitting FRAP/Spt experiments and be cautious about their interpretation. For example, interpreting a kinetic behavior as a DNA-specific binding event can be accurate, only if backed up with measurements with a mutant version of the DNA binding domain.

We apologise if some of this information was not evident. The number of trajectories is provided in new Figure S1F, which indicates the number of trajectories analyzed for each condition in Figure 1.

We have now added also the numbers of trajectories analyzed for the ring experiments.

The comments on page 7 about bleaching refer to the technical limitations of the SPT approach. However, as bleached particles cannot be distinguished from those that leave the plane of imaging, they have not been filtered or removed. We have not sought to make claims about absolute residence times for that reason. Rather the point is to make a comparison between the different molecules. As the same fluorescent ligand and imaging conditions are used in all the experiments, all the samples are equivalently affected by bleaching. We subdivide trajectories according to their properties and infer that those which are essentially stationary are bound to chromatin, as is common practice in the field. We note that we have previously shown that a DNA binding mutant of CSL does not produce a hub at E(spl)-C in Notch-ON conditions and has a markedly more rapid recovery in FRAP experiments (Gomez-Lamarca et al, 2018) consistent with the slow recovery being related to DNA binding. This point has been added to the text (page 8).

(4) The authors should quantify their RNAi efficiency for Hairless-RNAi, Med13-RNAi, white-RNAi, yellow-RNAi, CBP-RNAi, and CDK8-RNAi.

We thank the reviewer for this comment. We have made sure that we are using well validated RNAis in all our experiments and have included the references in Table S2 where they have been used. We have now evaluated the knock-down in the precise conditions used in our experiments by quantitative RT-PCR and added those data, which show efficient knock-down is occurring, to new Supplementary Figure S1D and Figure S3J. We note also that the RNAi experiments are complemented by experiments inhibiting the complexes with specific drugs and that these yield similar results.

(5) Figure 3 A: could the author show that transcription is indeed inhibited upon triptolide treatment with smFISH (with for example m3 probes)? Why not use alpha-amanitin?

We thank the reviewer for this suggestion. We had omitted the smFISH data from this experiment in error. These data have now been added to new Supplementary Figure S3A and clearly show that transcription is inhibited following 1 hour exposure to triptolide. Triptolide is a very fast acting and very efficient inhibitor of transcription that acts at a very early step in transcription initiation. In our experience it is much more efficient than alpha-amanitin and is now the inhibitor of choice in many transcription studies.

(6) Figure 4 typo: panel B should be D and vice versa. Accessibility panels are referred to as Figure 4D, D' in the text but presented as panel B in the Figure.

We thank the reviewer for noting this mistake, it is now changed in the main text.

(7) The authors must add their optogenetic manipulation protocol to their methods section.

The method is described in detail in a recently published paper that reports its design and use. We have now also added a section explaining the paradigm in the methods (Page 31) as requested.

(8) Figure 3G needs a Y-axis label.

Our apologies, this has now been added.

(9) The authors should note why there was a change of control in Figure 3D compared to 3E and G (yellow RNAi vs white RNAi).

This is a pragmatic choice that relates to the chromosomal site of the RNAis being tested. Controls were chosen according to the chromosome that carries the UAS-RNAi: for the second chromosome this was yellow RNAi and for the third white RNAi. This is explained in the methods.

(10) Figure 1 would benefit from a diagram describing the genomic structure of the E(spl) locus and the relative position of the labelled locus within it.

We thank the reviewer for this suggestion and have added a diagram to Supplementary Figure S1A .

Reviewer #2 (Recommendations For The Authors):

Minor criticisms and typos:

Pet peeve: in some of the figure panels they are labeled Notch ON or OFF, but in others they are not, albeit that info is included in the figure legend. For the ease of the reader/reviewer, would it be possible to label all relevant figure panels either Notch ON or OFF for clarity?

We thank the reviewer for this suggestion and have modified the figures accordingly.

Page 7, top. "In comparison to their average distribution across the nucleus, both CSL and Mam trajectories were significantly enriched in a region of approximately 0.5 μm around the target locus in Notch-ON conditions, reflecting robust Notch dependant recruitment to this gene complex." Are the authors referring to Figure 1D here?

Thank you, this figure call-out has been added in the text.

Page 9. "...reported to interact with p300 and other factors (Figure S2B)." I believe the authors mean Figure S2C and not S2B.

Thank you, this has been corrected in the text.

Page 9. There is no Figure S2D.

Apologies, this was referring to Figure S1D, and is now corrected in the text.

Page 11: "...were at very reduced levels in nuclei co-expressing MamDN (Figure 4B).." Should be Figure 4CD.

Thank you, this has been corrected in the text.

Page 12: "...which was maintained in the presence of MamDN (Figure 4D, D')." Should be Figure 4B.

Thank you, this has been corrected in the text.

Reviewer #3 (Recommendations For The Authors):

In the Results section on Hub, the paragraph starting with "Third, we reasoned . ." the callout to Figure S2D should be Fig S1D.

Thank you, this has been corrected in the text

Figures: The font size in the Figures is so small that most words and numbers cannot be read on a printout. One has to go to the electronic version and increase the size to read it. This reviewer found that inconvenient and often annoying.

We apologise for this oversight, the font size has now been adjusted on all the graphs etc.

Figure legends: the legends are terse and in some cases leave explanations to the imagination (e.g. "px" in Figure 2E). It would be useful to go through them and make sure those who are not a Drosophila Notch person and not a transcription biochemist can make sense of them.

Our apologies for the lack of clarity in the legends. We have gone over them to make them more accessible and less succinct.

Version 2

Abstract

Summary

Developmental programming involves the accurate conversion of signaling levels and dynamics to transcriptional outputs. The transcriptional relay in the Notch pathway relies on nuclear complexes containing the coactivator Mastermind (Mam). By tracking these complexes in real time, we reveal that they promote formation of a dynamic transcription hub in Notch ON nuclei which concentrates key factors including the Mediator CDK module. The composition of the hub is labile and persists after Notch withdrawal conferring a memory that enables rapid reformation. Surprisingly, only a third of Notch ON hubs progress to a state with nascent transcription, that correlates with Polymerase II and core Mediator recruitment. This probability is increased by a second signal. The discovery that target-gene transcription is probabilistic has far-reaching implications because it implies that stochastic differences in Notch pathway output can arise downstream of receptor activation.

Introduction

Cells face the challenge of transmitting information accurately, so that cell-surface signals are translated into correct transcriptional responses, and how this is achieved mechanistically remains a major question. Notch is a key signalling pathway that leads to gene activation when ligand and receptor engage upon cell contact 1–3. The physical interaction brings about a conformational change that permits proteolytic cleavage and release of Notch Intra Cellular Domain (NICD) 4–6. This moiety forms a complex with CSL (CBF-1/RBPJ-κ in mammals, Suppressor of Hairless in Drosophila and LAG-1 in C. elegans), a transcription factor that binds to specific DNA motifs, and Mastermind (Mam) a co-activator 7–9. This tripartite activation complex promotes transcription from the target genes where it is recruited. The sites of recruitment differ according to the cellular context, resulting in different transcriptional outcomes and suggesting that other factors are important in preparing the targets for activation 10. In addition, release of NICD brings about rapid and robust transcriptional responses within minutes, raising the question how the molecules of cleaved Notch achieve this so efficiently 11–13.

Regulation of transcription must be tightly controlled in space, time, and genomic location 14,15. Many different studies report that sequence-specific transcription factors, key coactivators, and RNA polymerase II (Pol II) itself undergo dynamic clustering within a nucleus 16–19. Clustering appears to be mediated by a combination of specific structure-mediated interactions (e.g. DNA-binding, protein-protein interactions) and multivalent interactions among intrinsically disordered regions (IDRs) present in most transcription factors 20,21. In this way, transcription is regulated by the formation of functionally specialized local protein microenvironments or transcription “hubs” associated with target enhancers 22. In some cases, these have the properties of ‘condensates’ whose formation and dissolution has been explained by the process of phase separation 23. As the resulting assembly is non-stoichiometric, it may enable a small number of transcription factor molecules to drive productive transcription. Such a mechanism could thus explain how NICD, whose nuclear levels are frequently below the level of detection in vivo, can successfully promote robust-target gene transcription 24. Indeed, all members of the Notch activator complex contain unstructured regions that could contribute to the assembly of a hub.

The formation of the tripartite Notch activator complex involves a conserved helix in the N-terminal region of Mam proteins which is responsible for the direct interactions with CSL and NICD 8,9. The remainder of the large Mam proteins are poorly conserved and appear to be predominantly unstructured albeit to have potential roles in recruiting other cofactors 25,26. For example, there is evidence that human MAML1 interacts with the histone acetyl transferase CBP/p300, which is present at Notch regulated enhancers in genome-wide studies 26–28, and whose recruitment is implicated in activating some targets 29,30. The C-terminal portion of MAML1 is also suggested to recruit CDK8, the enzymatic core of the Mediator kinase module 31–33. What role these factors play in the recruitment, dynamics and assembly of functional Notch transcription assemblies, and whether these acquire hub-like properties is unclear.

Live imaging of endogenously tagged proteins offers a non-invasive approach to probe the assembly and composition of transcription hubs in vivo. Using this strategy, we have previously shown that CSL is recruited in a very dynamic manner to a target genomic locus in vivo 34. However, as CSL exists in both co-repressor and co-activator complexes 35, the extent that these dynamics reflect the characteristics of the activation complex remains to be established. Here we incorporated fluorescent tags into the endogenous Mam protein to investigate the behaviours of this Notch nuclear co-activator in vivo, in combination with a method for live-imaging of a target genomic locus that responds robustly to Notch activation. The emerging model is that Notch activity leads to the formation of ‘transcription hubs’ that exist in different states. When Mam is present, key components of the transcriptional machinery become locally concentrated but, surprisingly, in the absence of synergising factors, the conversion to productive transcription only occurs stochastically. In addition, the open chromatin state that is generated decays slowly after Notch withdrawal, providing a memory that enables a more rapid response to a subsequent round of Notch activation.

Results

Dynamics of Mastermind recruitment in relation to its partner CSL

The Mam co-activator is an integral part of the Notch transcription complex 8,9,30,36. To track Mam dynamic behaviours in vivo we inserted GFP or Halo into the N-terminus of the endogenous Mam using CRISPR-Cas9 genome editing 34. The resulting flies are homozygous viable with no evident phenotypes, indicating that the tagged Mam proteins are fully functional. We first set out to compare the recruitment and dynamics of CSL and Mam at a target locus in Notch OFF and Notch ON conditions, taking advantage of Drosophila salivary glands, where the polytene (multiple copy) chromosomes aid detection of chromatin-associated complexes. We used the Int/ParB system, where fluorescently labelled ParB proteins bind to inserted Int sequences, to detect the well-characterised Enhancer of split complex [E(spl)-C], which contains multiple Notch-regulated genes (Figure S1A, 34). Because of the aligned copies of the genome in the polytene chromosomes, the target genomic locus appears as an easily distinguishable fluorescent “band” in each nucleus during live imaging (Figure 1A).

Enrichment and dynamics of Mastermind at E(spl)-C.

(A) Schematic overview of live imaging system used. Salivary glands from Drosophila larvae (left) have large nuclei with polytene chromosomes (centre, grey shading) in which E(spl)-C locus is detected as a band (centre, red) by live imaging when labelled using Int (orange)/ParB (red) system (right and Figure S1A). Recruitment of activation complexes, [CSL (green), NICD (purple) and Mastermind (magenta)] and of co-repression complexes [CSL and Hairless (brown)] is measured by their colocalization with E(spl)-C.

(B) Live-imaging of GFP::Mam and GFP::CSL as indicated in relation to E(spl)-C marked by Int-ParB (magenta) in nuclei from Notch OFF (1151Gal4; UAS-LacZ) and Notch ON (1151Gal4; UAS-NΔECD) salivary glands. CSL and Mam are enriched at E(spl)-C in Notch ON but not Notch OFF cells. Average enrichment: each pixel represents average enrichment of all aligned images, centred on E(spl)-C locus (0, grey bar). Enrichment profile: mean enrichment, with SEM, plotted on x-axis relative to position, y axis, centred on E(spl)-C (0). Grey area indicates region used for max enrichment. Max enrichment: mean of 10 pixels centred on E(spl)-C (Figure S1B) (Mam OFF, n=32, Mam OFF, n = 30; CSL OFF, n = 45, CSL ON, n = 28, For p-values see Table S4). Box encompasses range between 0.25 and 0.75 quantile, whiskers extend to furthest points not considered outliers, bar marks median, cross marks mean and each dot is the value for one nucleus. Scale bars represent 5um. Full genotypes for all figures are provided in Table S3.

(C, C’) Dynamics of CSL::GFP (green), GFP::Mam (magenta) and Hairless::GFP (brown) at E(spl)-C in Notch ON cells measured by FRAP. Recovery of the indicated proteins was measured in a point bleached region-of-interest centred on E(spl)-C and normalised by using another region and efficiency of bleaching. Dots indicate 50% recovery. Legend summarizes numbers of nuclei (n) and time to 50% recovery (t1/2). Error represents the standard error of the mean (SEM). (C’) FRAP analysis of CSL and Mam in cells depleted for Hairless (1151Gal4; UAS-Hairless RNAi) controls from C are included for comparison.

(D) Trajectory density at E(spl)-C relative to whole nucleus from SPT of CSL::Halo (green) and Halo::Mam (pink) in Notch OFF and Notch ON cells. Both CSL and Mam are significantly enriched in Notch ON. Box plots as in (B). CSL-OFF, n = 5, CSL-OFF, n = 7; Mam-OFF, n = 5, Mam-ON, n = 13; For number of trajectories see Figure S1F and for p-values see Table S4.

(E, E’) Survival curves depicting duration of trajectories in SPT of the indicated Halo fusion proteins in whole nuclei (E) and at E(spl)-C (E’), in Notch ON cells. Whole nuclei H2B::Halo trajectories are included for comparison in both graphs. For number of trajectories, see Figure S1F. Error bars represent 95% confidence intervals, obtained from bootstrapping with 100 resampled datasets 103.

The Notch pathway is normally inactive in salivary glands, providing a baseline Notch OFF condition. This was converted to Notch ON by the expression of a constitutively active form of Notch, NΔECD, using the GAL4-UAS system which allows tissue-specific and temporal control. Because it lacks the extracellular domain, NΔECD is constitutively cleaved by gamma-secretase to release NICD, mimicking ligand induced activation 37–39. Comparing the localization of GFP::Mam in Notch OFF and Notch ON conditions, it was immediately evident that Mam was robustly recruited to E(spl)-C in Notch ON conditions in a similar manner to CSL 34. In Notch OFF conditions, both proteins were diffuse throughout the nucleus, with a low level of CSL, but not Mam, present at E(spl)-C. In Notch ON conditions, strong enrichment of both Mam and CSL was consistently detected around E(spl)-C, where the two proteins co-localise in a correlated manner (Figure 1B, S1C).

To compare the dynamics of Mam and CSL at E(spl)-C in Notch ON conditions, we performed Fluorescence Recovery After Photobleaching (FRAP) focussed on the region defined by the locus-tag. Unlike CSL, which had a rapid recovery (t1/2=9 secs), Mam exhibited slower dynamics (t1/2=40 secs) and failed to fully recover over the time-course of the experiment (Figure 1C). Slower recovery can arise from higher proportion of bound molecules, longer residence times and/or slower diffusion coefficients40. The results therefore suggest that the Mam-containing activation complexes have different properties from the majority of CSL complexes.

The other main partner for CSL in Drosophila is the co-repressor Hairless. In FRAP experiments Hairless has a fast recovery, with a profile close to that of CSL (Figure 1C), consistent with a significant fraction of CSL being complexed with Hairless even in Notch ON conditions34. The difference in dynamics between CSL and Mam could therefore be explained by the former being involved in two different complexes with different dynamics. To test this, we depleted Hairless using RNAi-mediated knock-down, which we validated by RT-qPCR, and measured the effects on enrichment and dynamics of CSL and Mam (Fig. S1D). As expected, Mam recruitment levels and the dynamics measured by FRAP were unchanged (Figure 1C’, S1E). In contrast, CSL levels and recruitment were reduced and its FRAP recovery was slowed, albeit not to the extent that it recapitulated the Mam profile (Figure 1C’, S1E). Together the results indicate that the activation complexes, containing CSL and Mam, have slower dynamics than the repressor complexes, containing CSL and Hairless, and that the recovery of CSL reflects its participation in the two types of complexes.

To further investigate the dynamics of Mam and CSL complexes, we performed single particle tracking (SPT) by sparse labelling of endogenous Halo::Mam and Halo::CSL in live tissue 41,42. Gaussian fitting-based localisation and multiple hypothesis tracking were used for detection and tracking of single particles within the nucleus with a ∼20nm precision34. Using a Bayesian treatment of Hidden Markov Models, vbSPT 43, trajectories were assigned into 2 states, defined by a Brownian motion diffusion coefficient, that correspond to “bound” chromatin associated molecules (diffusion coefficient 0.01µm2/s) versus more freely diffusing complexes (diffusion coefficient >0.25µm2/s). A greater proportion (55%) of Mam complexes were in the bound state than CSL complexes (39%), consistent with the differences between their FRAP curves (Figure S1F). We also analysed the density of particle trajectories in relation to the E(spl)-C locus in Notch OFF and Notch ON conditions. In comparison to their average distribution across the nucleus, both CSL and Mam trajectories were significantly enriched in a region of approximately 0.5 µm around the target locus in Notch ON conditions, reflecting robust Notch dependant recruitment to this gene complex (Figure 1D).

To assess whether Mam complexes have longer residence times once recruited to the chromatin, we analysed the duration of trajectories for Mam, CSL and Hairless. Long trajectories correlate to bound complexes, because faster moving particles are rapidly lost from the field of view, and the length of time they are detectable is an indication of relative residence times. There were clear differences between the trajectory durations for Mam, CSL and Hairless. Mam trajectories had the longest durations of up to 15 secs, Hairless trajectories were the shortest (up to 5-7 secs) and CSL trajectories were intermediate (up to 10 secs) (Figure 1E). The differences were recapitulated when only the trajectories in the region around E(spl)-C were analysed (Figure 1E’). The residences are likely an underestimation because bleaching and other technical limitations also affect the track durations44. Nevertheless, these data confirm that Mam-containing complexes have on average longer residence times than other CSL complexes which, together with the higher proportion of bound molecules overall, explains the slower recovery dynamics measured by FRAP.

The fact that CSL dynamics, measured by FRAP and SPT, are intermediate between Hairless and Mam fit well with it being present in two types of complexes [co-activator (Mam) and co-repressor (Hairless)]. However, whether the contribution from CSL co-repressor complexes can fully explain all the observed differences between CSL and Mam is not fully clear. First, when we depleted Hairless, so that the majority of CSL present would be in co-activator complexes, CSL FRAP recovery curves were still substantially different from those of Mam. Second, none of the CSL trajectories had a duration approximating those of the longest-lasting Mam trajectories, despite that over 50,000 CSL trajectories were tracked (compared to 14,000 Mam, Figure S1F). It is possible therefore that, once recruited, Mam can be retained at target loci independently of CSL by interactions with other factors so that it resides for longer.

Hub-like properties of CSL-Mastermind complexes in Notch active cells

The enrichment of CSL and Mam around E(spl)-C that we detect by live-imaging is not unexpected because this locus has multiple genes containing CSL binding-motifs 34,45. However, the diffuse enrichment was not maintained when the tissues were fixed, a characteristic reported for proteins present in condensate-like hubs 46 (Figure S2A). The localized concentration of exchanging CSL and Mam complexes around E(spl)-C in Notch ON nuclei may therefore have properties of a transcription hub, with some of the recruitment being reliant on weak interactions mediated by low-complexity regions 47–49.

First, we asked to what extent CSL recruitment was correlated with the number of CSL binding motifs. To do so, we took advantage of fly strains containing multiple copies of CSL motifs inserted at an ectopic position in the genome 50 and compared the recruitment with insertions containing 12 or 48 CSL motifs. In Notch ON conditions, these insertions were sufficient to generate an ectopic band of CSL recruitment similar to the native E(spl)-C (Figure 2A). Remarkably the amounts of CSL recruited to loci with 12 and 48 ectopic sites were almost identical. Thus, there is not a direct correlation between the number of motifs and the amount of CSL recruited under these conditions, although we note that DNA binding is a pre-requisite as no recruitment occurs with CSL mutant that lacks DNA binding 34. Nor does it appear that the amounts of CSL complexes are a limiting factor as there was no decrease in recruitment at the endogenous E(spl)-C locus, even in nuclei with a 48 CSL-motif insertion.

Hub-like properties of recruited complexes in Notch ON nuclei.

(A) Representative images of GFP::CSL in Notch ON nuclei containing an ectopic array of 12 or 48 CSL binding sites. GFP::CSL is recruited to E(spl)-C (yellow) and the ectopic array (blue). Average enrichment, Enrichment profile as in Figure 1B; (ectopic array 12, n=45, ectopic array 48, n = 40; E(spl)-C 12, n=45, E(spl)-C 48, n=40).

(B) Domain organisation of the indicated proteins is diagrammed above the prediction scores (Metapredict V2) of protein disorder for each. Regions tested as IDRs are indicated.

(C) Average enrichments and enrichment profiles of NICD-PEST IDR (Notch OFF, n = 62, Notch ON, n = 67) and CSL-nt IDR (Wilcoxon rank sum test: Notch OFF, n = 8, Notch ON, n = 40) at E(spl)-C in Notch OFF and Notch ON cells.

(D) Average enrichments and enrichment profiles (as in Figure 1B) of GFP::Mam at E(spl)-C in nuclei expressing intact (Nact) and IDR deleted (Nact ΔIDR) Notch constructs (Nact, n = 58, Nact ΔIDR, n = 55). p-values in Table S4

(E) Diagram illustrates concentric ring analysis of single particle trajectories at increasing distances from E(spl)-C. Graphs plot proportions of bound (dark shading) and diffusive (light shading) particles of Halo::Mam or Halo::CSL within each ring. Dashed lines indicate nuclear proportions of each population as indicated. For number of trajectories see Figure S1F.

See also Supplementary Figure S2.

Second, we questioned whether the non-stochiometric recruitment of activator complexes to loci with CSL motifs might involve additional weak protein-protein interactions between some components, as observed in several transcription hubs where intrinsically disordered protein domains play a part 20,51. We used Intrinsically Disordered Region (IDR) prediction algorithms 52 to identify IDRs within NICD, CSL, Mam and Mediator1 (Med1, part of the mediator complex) (Figure 2B) and generated transgenic flies where each IDR was expressed as a fusion with GFP. The recruitment of the IDR::GFP fusions to E(spl)-C was then measured in Notch ON and OFF conditions (Figure 2C). Of the IDRs tested, the C-terminal region from NICD was the most strongly enriched at E(spl)-C in Notch ON conditions. A low level of enrichment was also evident for a shorter NICD fragment, that lacked the C-terminal PEST sequences and for Med 1 IDR but not for the terminal regions of CSL. Surprisingly, little or no enrichment occurred with Mam-IDRs, despite that this large protein has been reported to interact with p300 and other factors (Figure S2C) and that the mammalian MamL1 forms puncta when over expressed 53. Notably however, Mam-nIDR::GFP fusion was present in droplets, suggesting it can self-associate when present in a high local concentration (Figure S2B, 54). Our results suggest therefore that the IDR in NICD may contribute to the localized enrichments at target loci in Notch ON cells, potentially in combination with IDRs present in other recruited factors. In support of this hypothesis, deletion of the IDR from NICD led to a reduction in the levels and stability of Mam recruitment to E(spl)-C (Figure 2D).

Third, we reasoned that the presence of a transcription hub, where complexes are retained in the vicinity via protein interactions, should result in local changes in the behaviours of CSL and Mam. After segregating the SPT trajectories according to their diffusion properties as described above (Figure S1F), we analysed the spatial distribution of the slow and fast populations in relation to the E(spl)-C, defined as the area within 550nm of the locus-tag. Based on the shape and centre of this region, concentric zones were defined at 550nm distances and the proportions of slow- and fast-moving particles in each zone were calculated. The results show that there is an enrichment for slow-diffusing and a depletion of fast-moving particles close to the locus, with these altered properties extending to a region of up to 1μm away (Figure 2E).

Together our data support the model that CSL-Mam complexes are recruited and form a hub of high protein concentrations around the target locus in Notch ON conditions and suggest that IDR interactions, as well as DNA binding, contribute to their recruitment and retention in a region surrounding the active enhancers.

Mediator CDK module is required for stable Mastermind recruitment

The hub-like properties and slower turnover of Mam complexes compared to CSL suggest that other factors will be involved in their stabilization. We therefore tested the consequences of inhibiting or depleting different factors to distinguish those required for Mam enrichment at E(spl)-C in Notch ON nuclei. We first asked whether active transcription was required for Mam recruitment, by exposing the tissue to triptolide, a fast and specific inhibitor of transcription initiation which effectively inhibited transcription in Notch ON cells (Figure S3A, 55). No change in Mam enrichment or recovery was detected, arguing that Mam recruitment is not dependent on initiation or RNA production (Figure 3A, B). As previous studies have reported an interaction between Mam and the histone acetylase CBP/p300 30,33,56,57, we next inhibited CBP activity using a potent and selective inhibitor A485 58. Tissues were exposed to A485 for one hour which was sufficient to severely reduce the levels of H3K27Ac and E(spl)m3 transcription, indicating that the treatment was effective (Figure S3B-C). Surprisingly however, there was no change in the recruitment of Mam in these conditions (Figure 3C). We also depleted CBP by RNAi and, as with the drug treatment, Mam recruitment was unchanged (Figure 3D, S3J). Based on these results, we conclude that CBP is not essential for the recruitment of Mam complexes to the hub formed at the E(spl)-CC locus.

Mam enrichment requires Mediator CDK module but is independent of transcription and CBP/p300.

(A-B) GFP::Mam recruitment levels and dynamics at E(spl)-C in Notch ON tissues treated with triptolide or DMSO as a control. (A) Average and Max enrichment as in Figure 1B, (DMSO n = 49, triptolide, n = 36). (B) FRAP recovery curves, plotted as in Figure 1C.

(C–F) Average enrichment and Max enrichment of GFP::Mam and GFP::CSL, as indicated, at E(spl)-C in Notch ON control and treated tissues. (C-D) No change in recruitment following inhibition (C, A485) or genetic knockdown (D, RNAi) of CBP/p300 compared to control, DMSO and yellow RNAi (yRi) respectively. (DMSO, n = 47, A485, n = 62; y RNAi, n = 26, CBP/p300 RNAi, n = 31). (E, F) Reduced recruitment of GFP::Mam (magenta) but not CSL::GFP (green) following (E) genetic knockdown of Med13 and (F) inhibition of CDK8 with Senexin B. (E: Mam control, n = 30, Mam Med13Ri, n = 45; CSL control, n=92, CSL Med13Ri, n = 36) and (F: Mam ctrl n = 27, Mam Senexin, n = 43; CSL ctrl, n = 34, CSL Senexin, n = 37).

(G) Genetic knockdown of Cdk8 results in decreased Mam recruitment compared to controls (ctrl, n=33, Cdk8 Ri, n = 32). Box plots in A-F as in Figure 1B. For p-values see Table S4.

(H) Average enrichment and max enrichment of Med13::YFP at E(spl)-C in Notch OFF, n = 28 and Notch ON, n = 31; For p-values see Table S4.

See also Supplementary Figure S3.

As several studies have suggested that the Mediator CDK module, containing Med12, Med 13, CDK8 and CycC, influences Notch dependent transcription, we next focused on the role of this complex 31,32,59. First, we depleted Med13 and CDK8 using well validated RNAi lines (60,61, S3J). Med13 depletion led to a loss of E(spl)m3 transcription, confirming its role (Figure S3E). In these conditions we detected a substantial reduction in Mam recruitment to E(spl)-C in Notch ON cells, suggesting the CDK module plays a role in retaining Mam at target sites (Figure 3E, G). Second, we incubated the tissues in Senexin B or Senexin A, two drugs that target CDK8 activity 62,63 for 1 hour prior to imaging. In both cases the treatment was sufficient to significantly impair Mam levels at E(spl)-C (Figure 3F, S3D) while a CDK9 inhibitor, NVP2, had no effect in the same assays despite a loss of E(spl)m3 transcription (Figure S3H-I). We note that human MamL1 is a high confidence CDK8 target 64, but as the phospho-sites are not conserved it is unclear whether Mam would also be a direct target of the kinase. Third we investigated whether CDK module was recruited to E(spl)-C, in Notch ON nuclei using an existing line in which YFP is fused to Med13 65,66. Significant enrichment of Med13::YFP was detected at E(spl)-C in Notch ON nuclei, (Figure 3H) demonstrating that it is present in the hub with Mam.

Despite their effects on Mam recruitment, neither depletion of Med13 nor Senexin treatments caused any significant reduction in the levels of CSL recruited. On the contrary, a small increase occurred (Figure 3E, F). One explanation for the differences in the effects on CSL and Mam could be that there is increased recruitment of CSL complexes containing the co-repressor, Hairless, following Senexin treatment. We therefore monitored Hairless recruitment under these conditions but detected no increase in Senexin treated nuclei (Figure S3F). However, the overall nuclear levels of Hairless are quite high making it difficult to detect and quantify any changes at E(spl)-C. We therefore investigated the impact of Senexin on recruitment of a mutant CSL (CSL-Hmut) with compromised Hairless binding. This retains normal recruitment in control Notch ON conditions (Figure S3G) but, strikingly, it is not enriched at E(spl)-C in Senexin treated tissues. This result suggests that CSL requires Hairless interaction for its recruitment to E(spl)-C upon CDK8 inhibition. The differences in the effects on Mam and CSL imply that the CDK module is specifically involved in retaining Mam in the hub, and that in its absence other CSL complexes containing Hairless “win-out”, either because the altered conditions favour them and/or because they are the more abundant.

Mastermind is not essential for chromatin accessibility

The observed differences between CSL and Mam in their dynamics and dependency on the CDK module led us to speculate that they make different functional contributions to the transcription hub. To probe the role of Mam, we examined the consequences when its recruitment to the complex was prevented. The N-terminal peptide from Mam functions as a dominant negative (MamDN) by occupying the groove formed by CSL-NICD and, when overexpressed, prevents transcription of target genes 8,9,67. As predicted, MamDN expression in Notch ON conditions prevented recruitment of full-length Mam to E(spl)-C (Figure 4A). To verify that MamDN also inhibited transcription under these conditions, we used single molecule Fluorescent in situ hybridization (smFISH) with probes targeting E(spl)-C transcripts that are robustly up-regulated in this tissue 34. Nascent transcripts of E(spl)m3 were detected at E(spl)-C in Notch ON nuclei and were at very reduced levels in nuclei co-expressing MamDN (Figure 4D, D’) confirming that target-gene transcription is blocked.

Mam is not necessary for CSL enrichment and chromatin opening but necessary for transcription and Mediator 13 enrichment.

(A-A’) Recruitment of GFP::Mam in Notch ON nuclei is perturbed by MamDN (1151-Gal4 UAS-MamDN) (A) while recruitment of GFP::CSL is unaffected (A’). Control UAS was included in Notch ON combinations, details of genotypes are provided in Table S3. Scale bars represent 5um.

(B-B’) Accessibility of Notch regulated enhancers adjacent to E(spl) -mβ and E(spl) -m3 under the conditions indicated was probed using ATAC-qPCR. Values were normalised to Rab11, error bars represent SEM. Closed control is a noncoding genomic region in chromosome 3 and open control’ Eip78, an ecdysone responsive region active in L3 larval stage.

(C) Average enrichment and max enrichment of Med13::YFP at E(spl)-C in Notch ON conditions only, and in combination with MamDN. Quantifications as in Figure 1B, OFF, n=28, ON, n = 31. For p-values see Table S4

(D-D’) Expression of E(spl)-m3, detected by smFISH in the conditions indicated; representative images from cytoplasm (upper; 1.8mm3) and nucleus (lower; centred on E(spl)-C, blue arrows). Scale bars represent 5um. Graphs (D’), number of RNA puncta (cytoplasmic, n = 18, 20 and 22) and locus intensity (nucleus, n = 6, 7 and 6) in Notch OFF, Notch ON ctrl and Notch ON MamDN as indicated, boxplots as described in Figure 1B.

See also Supplementary Figure S4.

By contrast, recruitment of CSL was unaffected by MamDN expression. It was robustly recruited to E(spl)-C at a similar level to untreated Notch ON nuclei despite the absence of full-length Mam (Figure 4A’, 34). Identical results were obtained when Mam was depleted by RNAi (Figure S4A, 34). The fact that CSL was still strongly enriched under these conditions argues that some of the Notch induced changes at E(spl)-C occur independently of the co-activator, as proposed previously 34. In contrast to CSL, the enrichment of Med13 was lost in the presence of MamDN revealing that Med13 recruitment requires Mam, as well as the converse (Figure 4C).

Since MamDN does not prevent CSL recruitment, although it blocks recruitment of Med13 and transcription, we hypothesised that some Notch-induced effects at target enhancers may not require Mam. One consequence from Notch activation is an increase in chromatin accessibility at sites where CSL is recruited 34,68. We therefore asked whether changes in chromatin accessibility still occur in the presence of MamDN by performing ATAC and analysing by qPCR the accessibility of the regulatory regions abutting E(spl)mβ and E(spl)m3 34. Both regions significantly increased in accessibility in Notch ON nuclei which was maintained in the presence of MamDN (Figure 4B). In addition, we found that the levels of GFP::NICD-IDR recruited to E(spl)-C were similar in the presence and absence of MamDN (Figure S4B), arguing that a modified hub is still formed.

Altogether these observations support the model that the increased chromatin accessibility elicited by Notch activation can occur independently of Mam and must rely on other functions conferred by CSL and NICD 34, while Mam is essential to recruit the Mediator CDK module and enable transcription.

CSL recruitment and chromatin accessibility persist after Notch inactivation, conferring memory

Our results suggest that there are two or more separable steps involved in forming an active transcription hub in Notch ON cells: a Mam-independent change in chromatin accessibility, a Mam-dependent recruitment of Mediator and initiation of transcription. If these are discrete steps, we reasoned that they might decay with different kinetics when Notch activity is removed. We took advantage of the thermosensitive Gal4/Gal80ts system to switch off Notch activity and assessed the consequences on CSL and Mam recruitment at two different time points: 4 hours and 8 hours after the switch-off. Imaging Mam::GFP and CSL::mCherry simultaneously, it was evident that Mam recruitment levels decreased more rapidly. CSL remained relatively constant through both time points while, in contrast, the levels of Mam at E(spl)-C decreased sharply after 4 hours (Figure 5A). Based on these results we propose that, after Notch activity decays, the locus remains accessible because when Mam-containing complexes are lost they are replaced by other CSL complexes (e.g. co-repressor complexes).

Effects of Notch withdrawal on hub composition imply memory state.

(A) Temperature paradigm: Withdrawal of Notch activity using the thermosensitive Gal4/Gal80ts system; after a 24h ON period, larvae were transferred to 18C (Gal4 inhibited) for the indicated periods (top) and enrichment of mCherry::CSL and GFP::Mam at E(spl)-C imaged. Average enrichment, enrichment profiles and max enrichment quantified as in Figure 1B. 24 hrs ON, n = 28, 24 hrs ON, 4 hrs OFF, n = 21, 24 hrs ON, 8 hrs OFF, n = 19. p-values are summarized in Table S4.

(B) Accessibility of enhancers at E(spl) -mβ and E(spl) -m3, in Notch ON and after Notch withdrawal (8 hours) was probed using ATAC-qPCR, as in Figure 4B.

(C) Recruitment of CSL::GFP and expression of E(spl)-m3, detected by smFISH in the conditions indicated. Robust recruitment of CSL::GFP at E(spl)-C persists after switching to none-permissive temperature (Notch OFF) for 24hours, representative image and graph (green) with intensity quantifications for multiple nuclei (24 hrs ON, n = 28, + 24 hrs OFF, n = 21). For smFISH, representative images from cytoplasm (upper; 1.8 mm3) and nucleus (lower; centred on E(spl)-C, blue arrow), are shown, scale bars represent 5um. RNA puncta are absent after 24 hours Notch OFF. Graphs show number of cytoplasmic RNA puncta (left) from n = 30, 16 regions and RNA fluorescence intensity at E(spl)-C in the conditions indicated from n = 10, 5 nuclei, with boxplots as described before (Figure 1B). p-values in Table S4.

(D) Optogenetic paradigm used for smFISH: Conditions used to switch ON and OFF Notch activity using

OptiC Notch 69

(E-F) Expression of E(spl)-m3, detected by smFISH in the conditions indicated. (E) Representative images from the nucleus (centred on E(spl)-C, blue arrows). Scale bars represent 5um. (F) Graphs for fluorescence intensity at E(spl)-C in the conditions indicated from n = 38, 21 and 43 nuclei, with boxplots as described before (Figure 1B). p-values in Table S4.

(G) Optogenetic paradigm used for live imaging: Conditions used to switch ON and OFF Notch activity using OptiC Notch 69.

(H) Recruitment of CSL::GFP and Mam::Halo at E(spl)-C, measured following blue light activation in OptiC Notch expressing tissues for 2 hours (ON) and after 4 hours in dark (OFF). GFP::CSL remains enriched, Halo::Mam is depleted.

(I). Recruitment of Halo::Mam in preactivated nuclei compared to naïve nuclei; temporal profile of Mam::Halo enrichment (right) and time taken to max enrichment (left). Error bars in the profile represents the SEM, values were normalised from 0 to 1 (similar to FRAP recovery, see Figure 1 and methods), n indicates the number of nuclei from >3 salivary glands. Boxplots parameters as in Figure 1B; naïve, n = 11, preactivated n = 9.

As Notch removal leads to a loss of Mam, but not CSL, from the hub, we hypothesised it would recapitulate the effects of MamDN on chromatin accessibility and transcription of targets. We therefore measured the accessibility of target enhancers at E(spl)mβ and E(spl)m3 at the 8-hour timepoint by ATAC. Neither enhancer exhibited any reduction in accessibility at this timepoint, consistent with the continued recruitment of CSL and the results obtained with MamDN (Figure 5B). In contrast, the gradual loss of Mam complexes was accompanied by reduced transcription, as detected by smFISH. Expression of E(spl)m3 was already significantly reduced at 4 hours, when levels of nascent transcripts at E(spl)-C and of cytoplasmic transcripts had both decreased (Figure S5A). By 8 hours both nascent and cytoplasmic E(spl)m3 transcripts were almost undetectable (Figure S5A). Thus, the changes in transcription correlate well with the reduction in Mam recruitment whereas the chromatin accessibility persists in the absence of Mam. We next investigated whether CSL remained enriched at the locus for an extended period following Notch inactivation. Indeed, we were still able to detect a strong CSL enrichment 24 hours after transfer to the non-permissive temperature at time when, consistent with shorter OFF periods, there was no detectable transcription of E(spl)m3 (Figure 5C).

One possible consequence of the prolonged CSL enrichment in Notch OFF conditions is that target loci will retain “memory” of previous Notch activation that would make them more receptive to a subsequent exposure to Notch activity. To investigate, we took advantage of the temporal control provided by optogenetic release of NICD using OptIC-Notch{ω} 69. We compared the response to Notch activation in cells that had been ‘preactivated’ with blue light for 4 hours and subsequently kept in the dark for 24 hours, and ‘naïve’ cells, which had been kept solely in the dark and hence had no prior Notch activity (Figure 5D-F). We confirmed that cells had no residual transcription after being kept in dark, Notch inactive conditions for 24 hours (“OFF” Figure 5E, F). Strikingly, after 30 mins in blue light to activate Notch, cells that had been preactivated showed higher levels of E(spl)m3 transcription compared to naïve cells, indicating that the previous Notch exposure renders them more sensitive (Figure 5E,F). To investigate whether a previous activation also influences Mam recruitment, we tracked its recovery in a live imaging experiment. As a starting point we deployed a 4 hour Off period, after which Mam was fully depleted at E(spl)-C while CSL enrichment remained (Figure 5G, H). During the subsequent activation and imaging, recruitment of Mam occurred more rapidly in preactivated cells in comparison to naïve cells, suggesting the former are primed to rapidly reform an active hub containing Mam (Figure 5I).

Together, our data indicate that CSL recruitment and increased chromatin accessibility persist after Notch removal and after the loss of Mam-containing activation complexes. This persistent CSL confers a memory state that enables re-assembly of an activation hub and more rapid initiation of transcription in response to subsequent Notch activity.

Probability of transcription conferred by Mastermind

When analysing the smFISH data we noticed that, even in Notch ON conditions, a fraction of nuclei lacked foci of nascent transcription at E(spl)-C. Since Mam was present at E(spl)-C in all nuclei (Figure 6A, S6A), this led us to question whether the presence of Mam-complexes was sufficient to recruit downstream factors required for transcription initiation. We therefore investigated the extent that RNA Pol II was recruited to E(spl)-C in Notch ON cells, using endogenous GFP::Rbp3 16. When scanning all nuclei, it was evident that Rbp3 enrichment at E(spl)-C in Notch ON cells was highly variable. Robust enrichment was detected in a subset of nuclei but, in other cases, there was little/no enrichment above nuclear levels (Figure S6B). Performing a Gaussian population fitting on the data, 2 populations gave the best fit and, based on these, 36% of nuclei had significant enrichment and 64% had similar levels to Notch OFF (Figure 6B, S6B). This differs from Mam where all nuclei fall into a single population that has significant Mam enrichment at E(spl)-C (Figure 6A, S6A). The striking difference in the proportions of nuclei with Pol II and with Mam enrichment implies that there is a limiting step, which results in transcription initiation being probabilistic/stochastic in these conditions.

Recruitment of Pol II and Med1 in Notch ON is infrequent and is augmented by ecdysone.

(A) GFP::Mam enrichment at E(spl)-C in Notch OFF and Notch ON conditions as in Figure 1, represented with a violin plot with mean marked by a cross and median by a bar. Right graph: proportions of nuclei with significant enrichment (obtained by performing a Gaussian fit on the fluorescence intensity Images, see Supplementary Figure S5). Over 90% nuclei have significant enrichment (Notch OFF, n = 32, Notch ON, n = 30)

(B) Recruitment of GFP::Rbp3, a subunit of Pol II, at E(spl)-C in Notch OFF and Notch ON nuclei. Images: example where enrichment is detected. Graph: proportions of nuclei with significant enrichment as in A (Notch OFF, n = 37, Notch ON, n = 43). All scale bars represent 5um.

(C) Recruitment of GFP::Med1 at E(spl)-C in Notch OFF and Notch ON nuclei. Images: example where enrichment is detected. Graph: proportions of nuclei with significant enrichment as in A (Notch OFF and ON, n = 38, 62). All scale bars represent 5um.

(D) Transcriptional activity in Notch ON cells detected using live imaging of GFP::MCP recruited to MS2 loops in transcripts produced by E(spl)bHLHmβ (see methods). Images: representative examples of active (upper) and inactive (lower) nuclei. Graph: Proportion of active nuclei (Notch ON, n = 53)

(E) Recruitment of GFP::Med1, mCherry::Rbp1 and Halo::CSL. Med1 and Rbp1 recruitments are correlated. Images: example of co-recruitment of Med1, Rbp1 and CSL. Graph: Correlation plot of max enrichment per nucleus at E(spl)-C, (marked by Halo::CSL recruitment), n = 48.

(F) Pretreatment with ecdysone in Notch ON conditions increases proportions of nuclei with recruitment of GFP::Rbp3, GFP::Med1 and with active transcription foci (MS2/MCP intensity) compared to controls (EtOH). Proportions of nuclei with significant enrichment defined as in A. Rbp3 ctrl, ecdysone: n = 39, 36. Med1 ctrl, ecdysone: n = 34, 34. MS2 ON ctrl, ON ecdysone, OFF ecdysone: n = 52, 42, 22.

See also Supplementary Figure S5.

It has been suggested that the core Mediator complex, as distinct from the CDK module, has an important role in bridging between enhancer bound transcription complexes and initiation complexes at promoters 70–72. To investigate core Mediator recruitment we generated a GFP::Med1 fusion by CRISPR genome-editing. Using this endogenously tagged protein, which is homozygous viable, we found that GFP::Med1 became enriched at E(spl)-C with a similar probability to Rbp3 (Figure 6C). Indeed when we compared directly the enrichment of Pol II and Med1 in Notch ON nuclei, by using mCherry::Rbp1 16 to monitor Pol II, we observed a significant correlation between the two proteins (R2 = 0.69, Fig 6E). No such correlation with CSL was observed, in the same experiments (Figures S6E).

To verify that only a subset of nuclei was transcriptionally active we used a strain where MS2 loops have been inserted into E(spl)mβ using CRISPR-Cas9 engineering 11 and imaged transcription live in Notch ON nuclei. In the presence of MCP-GFP, which binds to MS2 loops in the RNAs produced, nascent sites of transcription appear as puncta that align into a band of fluorescence at E(spl)-C, due to the multiple aligned gene copies. A robust band of MCP/MS2 nascent transcription at E(spl)-C was detected in Notch ON conditions, demonstrating that some nuclei were actively transcribing (Figure 6D, S6D). However, this was present in only a third (37%) of nuclei. These data show that only a subset of nuclei engage in active transcription, and that the proportion of active nuclei is similar to that with recruitment of Pol II and Med1.

These data demonstrate that presence of Mam-complexes is not sufficient to reliably drive all the steps required for transcription in every Notch ON nucleus. Instead, it appears that transcription is initiated stochastically. Based on previous study in the Drosophila embryo, where transcription in responding nuclei was highly synchronised 73, this probablistic outcome was unexpected although stochastic transcriptional responses have been observed in C. elegans 74. We note that such properties can only be detected using in vivo imaging approaches to monitor the responses of individual nuclei, as we have done here.

Ecdysone cooperates with Notch to increase probability of transcription

Even though relatively few Notch ON nuclei became transcriptionally active in our experiments, they all had robust recruitment of Mam complexes and of Med13 at E(spl)-C. Thus, in many respects the gene locus becomes competent or poised for transcription in all nuclei. We wondered, therefore, whether the presence of a second stimulatory signal would increase the probability of loci becoming transcriptionally active. In normal development, salivary glands become exposed to the steroid hormone ecdysone a few hours after we perform our experiments. Two observations suggested that ecdysone was a good candidate for a cooperating signal. First, previous genome-wide studies detected ecdysone receptor binding in the E(spl)-C region 75. Second, we noticed in rare samples containing an older gland, that it had a higher proportion of active nuclei. We therefore exposed the early-stage Notch ON salivary glands to ecdysone and analysed the proportion of nuclei with MS2/MCP puncta. Strikingly, the proportion of active nuclei increased dramatically to 70% following ecdysone treatment. No such effects were seen in Notch OFF nuclei where E(spl)mβ remained silent even after ecdysone treatment (Figure 6F, S6H).

We further tested the effect of ecdysone by measuring changes in enrichment of Pol II and Med1 at E(spl)-C. The proportion of nuclei with clear enrichment of each factor was significantly increased (Figure 6F, S6F,G). Thus, the switch from the more stochastic transcription elicited by recruitment of Mam alone to the more robust initiation when ecdysone was added, correlated with the presence of Med1 and Pol II. As Mediator is reported to stabilize enhancer-promoter interactions, its recruitment may be what limits the probability of transcription 70,76.

Discussion

Understanding the mechanisms by which signaling levels and dynamics are accurately converted to transcriptional outputs is fundamental for developmental programming. Here we used live imaging approaches to probe how this occurs in the context of the Notch pathway, where the transcriptional relay relies on the single NICD released by each activated receptor and the nuclear complexes it forms with its partners CSL and Mam. By tracking these complexes in real time we unveiled three important features (Figure 7): (i) in Notch ON nuclei, the activation complexes promote formation of a dynamic protein hub at a regulated gene locus that concentrates key factors including Mediator CDK module, (ii) the composition of the hub is changeable and the footprint persists after Notch withdrawal conferring a memory that enables rapid reactivation, (iii) transcription is probabilistic in Notch ON nuclei, such that only a third of nuclei with a hub are actively transcribing. This has far-reaching implications because it reveals that stochastic differences in Notch pathway output can arise downstream of receptor activation.

Model illustrating different modes of Notch transcription-hub.

In the absence of Notch activity (Notch OFF) target genes are inactive, CSL is complexed with co-repressors (e.g. Hairless). Following Notch activation (Notch ON), released NICD (purple) generates a localized high concentration of transcription factors around target enhancer(s) referred to as a “hub” (pink). Open hub: CSL (green) recruitment and accessible chromatin can occur in the absence of Mam. Engaged hub: presence of Mam (magenta) favours recruitment of additional factors, including Mediator CDK8 module (orange). Active hub: productive transcription, transition to this mode, with core Mediator (orange) and Pol II (pale blue) enrichment, is stochastic (dotted arrow). The probability can be enhanced by secondary signal, such as provided by ecdysone. Memory hub: CSL enrichment and chromatin accessibility remain after withdrawal of NICD.

Notch transcription complexes form a hub

Many recent studies have demonstrated that hubs or condensates play key roles in gene-expression regulation by maintaining a high local concentration of transcription factors and other regulatory complexes 17,22,47,70. By tracking the co-activator Mam in real time, we show that Notch transcription complexes become enriched around a target locus in a zone that has characteristics of a hub. First, the high-concentration zones are highly dynamic and undergo a continual exchange between activator and repressor complexes. The former exhibit longer residence times, suggesting they are stabilised by other interactions. Second, recruitment of CSL-containing complexes is non-stoichiometric with respect to CSL-binding motifs and can in part be recapitulated by an IDR from NICD. Although deletion of the IDR biases against activation complex recruitment, our evidence suggests that individual IDRs make a minor contribution, as consistent with an unbiased study of IDRs 77. Third, the enriched zone is readily detectable by live imaging, but is sensitive to fixation, as has been reported for some subcellular protein condensates 46. Although transcription hubs have frequently been associated with phase separation, examples are emerging where high local concentrations of transcription factors can be achieved without it 21. Similarly, the locally enriched Notch-induced hubs in our experiments do not manifest clear properties of liquid condensates despite being non-stoichiometric.

We propose that NICD and Mam participate in multivalent interactions which increase the likelihood of the tripartite activation complexes remaining in the chromatin-associated state and which facilitate recruitment of co-activators to form a local hub (Figure 7). The latter include subunits of CDK module, Med 13. We find that perturbations to CDK module compromise Mam levels in the hub and vice versa, highlighting its importance. Indeed, the CDK module has been reported to interact directly with Mam and to play a role in Notch signalling output 31,32, and recent studies in mammalian cells indicate CDK8 acts positively in signal-induced gene expression 78. Although its recruitment is proposed to prepare genes for activation, the transition to activation necessitates CDK8 release, since CDK module sterically inhibits core Mediator-Pol II interactions76,79. This is consistent with our results showing that Med13 recruitment is highly dynamic and that, despite most nuclei showing Med13 enrichment within the hub, the probability of transcription is low, as discussed below.

Role of hub in conferring transcriptional memory

Assembly of the transcription hub at E(spl)-C depends on Notch activity but, surprisingly, can form in the absence of the co-activator Mam, albeit the composition differs. Notably, CSL complexes continue to be highly enriched and the chromatin at E(spl)-C enhancers is accessible in the absence of Mam (Figure 7). These properties also persist for many hours after withdrawal of Notch activity and in this way, enhancers that have been switched on by Notch retain an imprint that can influence their response to a subsequent signal, a phenomenon known as transcriptional memory80,81. As an indication, after a previous Notch activation event, the conversion to an activation hub with Mam enrichment occurs more rapidly than in the absence of a prior active state and there is enhanced transcriptional activity. Similar accelerated recruitment of STAT1 to promoters occurred in cells with IFNγ-induced transcriptional memory 82, and augmented transcription levels have been seen in several contexts 83,84.

The concept that enhancers retain a memory of a recent Notch signal has several implications. First, naïve and pre-activated loci will respond with different kinetics, which could bias cell-fate decisions when iterative Notch signals are deployed 85. Second, although our analysis is focused on post-mitotic cells, the footprint may be inherited by daughter cells as CSL is retained on mitotic chromosomes in at least some cell-types 86. If this is the case, it may explain why enhancer decommissioning is important to switch-off the Notch response in some contexts 87–89. Third, it could explain some of the observed recruitment of CSL co-repressor complexes to active chromatin if they are involved in sustaining the hub when the signal decays 90,91.

Transcription initiation is probabilistic

Because the in vivo live imaging enables us to monitor the responses of individual nuclei, we made the surprising discovery that the presence of Mam-containing hubs at E(spl)-C is not sufficient to guarantee transcription. Instead, our results reveal that Mam recruitment confers a 1 in 3 chance of transcription. As transcription initiation is inherently stochastic 92,93, it is not necessarily unexpected that only a fraction of nuclei would be active at any one time. Indeed, stochastic response to Notch activity has been detected in the C.elegans gonad, where a similarly probabilistic response dependent on Notch was detected 74,94. However this situation differs from the Drosophila mesectoderm, where the response to Notch was fully penetrant and highly synchronised across multiple nuclei, arguing that Notch regulated transcription is not a priori stochastic 73. Furthermore, it has been suggested that recruitment of the activation complex would be the limiting factor. In our context it is evident that recruitment of Mam activation complexes is not the limiting step. This also differs from experiments with the Glucocorticoid receptor, which concluded that it’s binding determined the frequency and length of the RNA bursts produced 95.

The observation that at any one time, only a third of nuclei are actively transcribing could be a reflection either of two cell populations, in which only a subset of cells transition to the ON state, or two dynamic states in which all nuclei transition between ON and OFF states. Monitoring transcription live for prolonged periods, we rarely detected transitions in the same nucleus, arguing that, if the differences reflect the dynamic states, both the ON and the OFF state must be of prolonged duration. At the same time, the live traces reveal that the transcription in the ON cells is dynamic, in keeping with frequent fluctuations in promoter states producing “bursts” 92,93. Thus we would argue that there are two probabilistic steps involved. The first is the transition of the target gene into an active ON state, which, in our basic Notch active condition, appears relatively infrequent, and the second is the switching ON and OFF of the promoter in ON cells to produce “bursts” of RNA, which is much more rapid. The presence of a second signal, in the form of ecdysone was sufficient to increase the probability of the first, slow, transition occurring.

In our experiments, the probability of nuclei being in the ON state correlated with the proportion of nuclei where Med1, a core Mediator subunit, was recruited into the hub at E(spl)-C. Mediator is an important intermediary between transcription factors and the preinitiation complex and it functions as an ‘integrator’ coordinating diverse and combinatorial inputs 70. The increased probability of Mediator recruitment can thus be explained by an increase in the valency of possible interactions when Notch and ecdysone are both active and present at the enhancers of target genes. We propose that this will be a general mechanism and that other signals/transcription factors will synergise with Notch complexes by adding to the valency of interactions and facilitating recruitment of coactivators to increase the probability of transcription. In different contexts this could toggle the response from one that is stochastic to one that is hard-wired.

The observation of a probabilistic transcriptional outcome downstream of Notch activity and Mam recruitment has profound implications because it has widely been assumed that stochastic differences in Notch pathway output arise due to fluctuations in ligands or ligand availability 96,97. Our results raise the possibility that differences may also arise due to intrinsic variations subsequent to/independent from receptor activation that affect the probability of transcription occurring. Furthermore, this probability can be modified by other signals, as we observed here with ecdysone, or by the presence of other transcription factors, as when enhancers are ‘primed’ 73 and/or cooperatively regulated 98–102.

Supporting information